## k-nearest neighbors

```{r, echo=FALSE, warning=FALSE, message=FALSE}
set.seed(2008)
library(tidyverse)
library(dslabs)
data("mnist_27")
# We use this function to plot the estimated conditional probabilities
plot_cond_prob <- function(p_hat=NULL){
  tmp <- mnist_27$true_p
  if(!is.null(p_hat)){
    tmp <- mutate(tmp, p=p_hat)
  }
  tmp %>% ggplot(aes(x_1, x_2, z=p, fill=p)) +
  geom_raster(show.legend = FALSE) +
  scale_fill_gradientn(colors=c("#F8766D","white","#00BFC4")) +
  stat_contour(breaks=c(0.5),color="black")
}
```

We introduced the kNN algorithm in Section \@ref(knn-cv-intro) and demonstrated how we use cross validation to pick $k$ in Section \@ref(caret-cv). Here we quickly review how we fit a kNN model using the __caret__ package. In Section \@ref(caret-cv) we introduced the following code to fit a kNN model:

```{r}
train_knn <- train(y ~ ., method = "knn", 
                   data = mnist_27$train,
                   tuneGrid = data.frame(k = seq(9, 71, 2)))
```

We saw that the parameter that maximized the estimated accuracy was:

```{r}
train_knn$bestTune
```

This model improves the accuracy over regression and logistic regression:

```{r}
confusionMatrix(predict(train_knn, mnist_27$test, type = "raw"),
                mnist_27$test$y)$overall["Accuracy"]
```

A plot of the estimated conditional probability shows that the kNN estimate is flexible enough and does indeed capture the shape of the true conditional probability. 

```{r best-knn-fit, echo=FALSE, out.width="100%"}
p1 <- plot_cond_prob() + ggtitle("True conditional probability")

p2 <- plot_cond_prob(predict(train_knn, newdata = mnist_27$true_p, type = "prob")[,2]) +
  ggtitle("kNN")

grid.arrange(p2, p1, nrow=1)
```

## Exercises 

1\. Earlier we used logistic regression to predict sex from height. Use kNN to do the same. Use the code described in this chapter to select the $F_1$ measure and plot it against $k$. Compare to the $F_1$ of about 0.6 we obtained with regression.

2\. Load the following dataset:

```{r, eval=FALSE}
data("tissue_gene_expression")
```

This dataset includes a matrix `x`: 

```{r, eval=FALSE}
dim(tissue_gene_expression$x)
```

with the gene expression measured on 500 genes for 189 biological samples representing seven different tissues. The tissue type is stored in `y`:

```{r, eval=FALSE}
table(tissue_gene_expression$y)
```

Split the data in training and test sets, then use kNN to predict tissue type and see what accuracy you obtain. Try it for  $k = 1, 3, \dots, 11$.

