--- 
title: "Introduction to Data Science"
author: "Rafael A. Irizarry"
date: "`r Sys.Date()`"
output: pdf_document
biblio-style: apalike
description: This book introduces concepts and skills that can help you tackle real-world
  data analysis challenges. It covers concepts from probability, statistical inference,
  linear regression and machine learning and helps you develop skills such as R programming,
  data wrangling with dplyr, data visualization with ggplot2, file organization with
  UNIX/Linux shell, version control with GitHub, and reproducible document preparation
  with R markdown.
documentclass: book
link-citations: yes
bibliography:
- book.bib
- packages.bib
site: bookdown::bookdown_site
always_allow_html: yes
---


```{r eval=FALSE, echo=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'), 'packages.bib')
```

# Preface {-}

These are the class notes used in the
 [HarvardX Data Science Series](https://www.edx.org/course/data-science-r-basics-harvardx-ph125-1x). The code to generate the notes is available on  [GitHub](https://github.com/rafalab/dsbook).

For updates follow [\@rafalab](https://twitter.com/rafalab)
 
This work is licensed under the [Creative Commons Attribution-NonCommercial-NoDerivs 3.0](http://creativecommons.org/licenses/by-nc-nd/3.0/us/) United States License.


## Is this book for you? {-}

This book teaches you how to analyze data using case studies. It has eight parts: __R Basics__, __Data Visualization__, __Probability__, __Statistical Inference and Modeling__, __Data Wrangling__, __Regression__, __Machine Learning__, and __Productivity Tools__. Each part has about 10 chapters, most with between 3-16 exercises. 

Here are four frequently asked questions that might help you determine if the book is for you.

1. I have been asked to teach a Data Science course, can this be used as a text book? Yes, this book is intended to be used as a text book in an introductory course for undergraduates. The book includes dozens of exercises and uses realistic datasets that can be used for problem sets. Depending on the level of your students you can teach the material in 1 or 2 semesters. 

2. I am a complete novice in data analysis and coding, is this book too advanced for me? Not necessarily. Although the Linear Regression and Machine Learning parts are at the level of an advanced undergraduate course, the book starts at a very basic level and gradually becomes more difficult. If you are a novice we recommend you  start with the R and RStudio installation sections in Productivity Tools, then try to complete the R Basics part. You can then make a decision keeping in mind that the next parts become increasingly more challenging. 

3. Does this book include everything I need to know about Data Science? This book focuses on data analysis. A good data analyst can definitely get a job as a data scientist but, often, organizations are looking to hire people with database management, software engineering, advanced programming, or software implementation skills that the book does not cover.

4. I am statistician and I already know R, do I need to read the entire book? No. You can probably skip most of the R Basics, Probability, Inference, and Regression parts. But you should consider reading the  _Case study_ chapters.

5. I only want to learn Machine Learning concepts, can I just skip to that part? Yes, but note that you will need to understand probability and statistical inference. Also, the examples in the Machine Learning part are implemented with R code.

6. I am particularly interested in learning how to use the __tidyverse__, does this book teach this? Absolutely! The Data Visualization and Data Wrangling Parts cover a substantial part of the __tidyverse__ including __ggplot2__. We use these packages throughout the book. 
In the first part of the book we start by teaching the very basics of R with the __tidyverse__ first covered in the Basic Data Wrangling Chapter. In the Regression Part we have chapters dedicated to more advanced packages such as __broom__. The Machine Learning and Productivity Tools parts are the least relevant to the tidyverse.

7. What do you mean by _Productivity Tools_? These are free tools we use to complete our data analyses in an organized way. Specifically these tools are R, RStudio, knitR, Unix, Git, and GitHub.

8. Do you make the solutions to the exercises available? Because the book is used in class, we do not make the solutions publicly available. However, we are working on a solution manual that we will make available to instructors. You can also contact the author and request to be added to the GitHub repository with the solutions.


# Acknowledgements {-}

This book is dedicated to all the people involved in building and maintaining R and the R packages we use in this book. A special thanks to the developers and maintainers of R base, the tidyverse and the caret package.

A special thanks to my tidyverse guru David Robinson. Also, many thanks to Stephanie Hicks who twice served as a co-instructor in my data science class and Yihui Xie who patiently put up with my many questions about bookdown. Thanks also to Karl Broman, from whom I borrowed ideas for the Data Visualization and Productivity Tools parts, and to Hector Corrada-Bravo, for advice on how to best teach Machine Learning. Thanks to Peter Aldhous from whom I borrowed ideas for the principles of data visualization section and Jenny Bryan for writing _Happy Git and GitHub for the useR_, which influenced our Git chapters. Thanks to 
Alyssa Frazee for helping create the homework problem that became the Recommendation Systems chapter. Also, many thanks to Jeff Leek, Roger Peng and Brian Caffo whose class inspired the way this book is divided and to Garrett Grolemund and Hadley Wickham for making the bookdown code for their R for Data Science book open. Finally, thanks to Alex Nones for proofreading the manuscript during its various stages.

This book was conceived during the teaching of several applied statistics courses, starting over fifteen years ago. The teaching assistants working with me throughout the years made important indirect contributions to this book. The latest iteration of this course is a HarvardX series coordinated by Heather Sternshein. We thank her for her contributions. We are also grateful to all the students whose questions and comments helped us improve the book. The courses were partially funded by NIH grant R25GM114818. We are very grateful to the National Institute of Health for its support.
A special thanks goes to all those that edited the book via GitHub pull requests: Huang Qiang (`nickyfoto`), Marc-André Désautels (`desautm`), Michail Schwab (`michaschwab`), Alvaro Larreategui (`alvarolarreategui`), Jake VanCampen (`jakevc`), and Guillermo Lengemann (`omerta`).






