--- 
title: "Introduction to Data Science"
author: "Rafael A. Irizarry"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
always_allow_html: yes
description: "This book introduces concepts and skills that can help you tackle real-world data analysis challenges. It covers concepts from probability, statistical inference, linear regression and machine learning and helps you develop skills such as R programming, data wrangling with dplyr, data visualization with ggplot2, file organization with UNIX/Linux shell, version control with GitHub, and reproducible document preparation with R markdown."
---


```{r eval=FALSE, echo=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'), 'packages.bib')
```

# Preface {-}

These are the class notes used in the
 [HarvardX Data Science Series](https://www.edx.org/course/data-science-r-basics-harvardx-ph125-1x). The code to generate the notes is available on  [GitHub](https://github.com/rafalab/dsbook).

We will be adding chapters during the next few months.

For updates follow [\@rafalab](https://twitter.com/rafalab)
 
This work is licensed under the [Creative Commons Attribution-NonCommercial-NoDerivs 3.0](http://creativecommons.org/licenses/by-nc-nd/3.0/us/) United States License.

# Acknowledgements {-}

This book is dedicated to all the people involved in building and maintaining R and the R packages we use in this book. A special thanks to the developers and maintainers of R base, the tidyverse and the caret package.

A special thanks to my tidyvserse guru David Robinson. Also, many thanks to to Stephanie Hicks who twice served as a co-instructor in my data science class and Yihui Xie who patiently put up with my many questions about bookdown. Thanks also to Karl Broman, from whom I borrowed ideas for the Data Visualization and Productivity Tools parts, and to Hector Corrada-Bravo, for advice on how to best teach Machine Learning. Thanks to Peter Aldhous from whom I borrowed ideas for the principles of data visualization section and Jenny Bryan for writing _Happy Git and GitHub for the useR_, which influenced our Git chapters. Thanks to 
Alyssa Frazee for helping create the homework problem that became the Recommendation Systmes chapter. Also, many thanks to Jeff Leek, Roger Peng and Brian Caffo whose class inspired the way this book is divided and to Garrett Grolemund and Hadley Wickham for making the bookdown code for their R for Data Science book open. Finally, thanks to Alex Nones for proofreading the manuscript during its various stages.

This book was conceived during the teaching of several applied statistics courses, starting over fifteen years ago. The teaching assistants working me throughout the years made important indirect contributions to this book. The latest iteration of this course is a HarvardX series coordinated by Heather Sternshein. We thank her for her contribution. We are also grateful to all the students whose questions and comments helped us improve the book. The courses were partially funded by NIH grant R25GM114818. We are very grateful to the National Institute of Health for its support.
A special thanks goes to all those that edited the book via GitHub pull requests: Huang Qiang (`nickyfoto`), Marc-André Désautels (`desautm`), Michail Schwab (`michaschwab`), Alvaro Larreategui (`alvarolarreategui`), and Jake VanCampen (`jakevc`).






