# Data types

We will be working with two types of variables: categorical and numeric. Each can be divided into two other groups: categorical can be ordinal or not, whereas numerical variables can be discrete or continuous.

When each entry in a vector comes from one of a small number of groups, we refer to the data as _categorical data_. Two simple examples are sex (male or female) and regions (Northeast, South, North Central, West). Some categorical data can be ordered, for example spiciness (mild, medium, hot), even if they are not numbers per se. In statistics textbooks, they sometimes refer to these as _ordinal_ data. 

Example of numerical data are population sizes, murder rates, and heights. Some numerical data can be treated as ordered categorical. We can further divide numerical data into continuous and discrete. Continuous variables are those that can take any value, such as heights, if measured with enough precision. For example, a pair of twins may be 68.12 and 68.11 inches respectively. Counts, such as population sizes, are discrete because they have to be round numbers.

Keep in mind that discrete numeric data can be considered ordinal. Although this is technically true, we usually reserve the term ordinal data for variables belonging to a small number of different groups, with each group having many members. In contrast, when we have many groups with few cases in each group, we typically refer to them as discrete numerical variables. So, for example, the number of packs of cigarettes a person smokes a day, rounded to the closest pack, would be considered ordinal, while the actual number of cigarettes would be considered a numerical variable. But indeed, there are examples that can be considered both numerical and ordinal when it comes to visualizing data.

## Exercise {-}

1. Define variables containing the heights of males and females like this:

```{r}
library(dslabs)
data(heights)
male <- heights$height[heights$sex=="Male"]
female <- heights$height[heights$sex=="Female"]
```

How many measurements do we have for each?


2. Suppose we can't make a plot and want to compare the distributions side by side. We can't just list all the numbers. Instead we will look at the percentiles. Create a five row table showing `female_percentiles` and `male_percentiles` with the 10th, 30th, 50th, ..., 90th percentiles for each sex. Then create a data frame with these two as columns.

    
3. Study the following boxplots showing population sizes by country:

    ```{r boxplot-exercise, echo=FALSE, message=FALSE}
library(tidyverse)
library(dslabs)
ds_theme_set()
data(gapminder)
tab <- gapminder %>% filter(year == 2010) %>% group_by(continent) %>% select(continent, population)  
tab %>% ggplot(aes(x=continent, y=population/10^6)) + geom_boxplot() + scale_y_continuous(trans = "log10", breaks = c(1,10,100,1000)) + ylab("Population in millions")
```

    Which continent has the country with the biggest population size ?
    
4. What continent has the largest median population size?
    
    
5. What is median population size for Africa to the nearest million? 
    
    
6. What proportion of countries in Europe have populations below 14 million?

    A. 0.99
    B. 0.75
    C. 0.50
    D. 0.25
  
    
7. If we use a log transformation, which continent shown above has the largest interquartile range ?
    
