# ggplot2 geometries {#other-geometries}

In Chapter \@ref(ggplot2) we introduced the __ggplot2__ package for data visualization. Here we demonstrate how to generate the types of plots shows in the previous chapter.

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(dslabs)
```

## Barplots

To generate a barplot we can use the `geom_bar` geometry. The defualt is to count the number of each category and draw a bar. Here is the plot for the regions of the US.

```{r barplot-geom}
murders %>% ggplot(aes(region)) + geom_bar()
```

We often already have a table with a distribution that we want to presnet as barplot. Here is an example such a table:

```{r}
data(murders)
tab <- murders %>% 
  count(region) %>% 
  mutate(proportion = n/sum(n))
tab
```

We no longer want `geom_bar` to count but rather just plot a bar to the height provided by the `proportion` variable. For this we need to provide `x` (the categories) and `y` (the values) and use the `stat="identity"` option. 

```{r}
tab %>% ggplot(aes(region, proportion)) + geom_bar(stat = "identity")
```

## Histograms 

To generate histograms we use the `geom_histogram`. By looking at the help file for this function, we learn that the only required argument is `x`, the variable for which we will construct a histogram. We dropped the `x` because we know it is the first argument.
The code looks like this:

```{r, eval=FALSE}
heights %>% 
  filter(sex == "Female") %>% 
  ggplot(aes(height)) + 
  geom_histogram()
```

If we run the code above, it  gives us a message:

> `stat_bin()` using `bins = 30`. Pick better value with
`binwidth`.
 
We previously used a bin size of 1 inch, so the code looks like this:

```{r, eval=FALSE}
heights %>% 
  filter(sex == "Female") %>% 
  ggplot(aes(height)) + 
  geom_histogram(binwidth = 1)
```

Finally, if for aesthetic reasons we want to add color, we use the arguments described in the help file. We also add labels and a title:

```{r height-histogram-geom}
heights %>% 
  filter(sex == "Female") %>% 
  ggplot(aes(height)) +
  geom_histogram(binwidth = 1, fill = "blue", col = "black") +
  xlab("Male heights in inches") + 
  ggtitle("Histogram")
```

## Density plots

To create a smooth density, we use the `geom_density`. Here the data previously shown as a histogram:

```{r, eval=FALSE}
heights %>% 
  filter(sex == "Female") %>%
  ggplot(aes(height)) +
  geom_density()
```

To fill in with color, we can use the `fill` argument.

```{r ggplot-density}
heights %>% 
  filter(sex == "Female") %>%
  ggplot(aes(height)) +
  geom_density(fill="blue")
```

To change the smoothness of the density we use the `adjust` argument to mulitply the default value by that `adjust`. For example, if we want the bandwidth to be twice as big we use:


```{r eval = FALSE}
heights %>% 
  filter(sex == "Female") + 
  geom_density(fill="blue", adjust = 2)
```


## Boxplot

The geometry for boxplot is `geom_boxplot`. As discussed, boxplots are useful for comparing distributions. For example, below are the previously shown heights for women, but compared to men. For this geometry, we need arugments `x` and the categories, and `y` the values. 

```{r female-male-boxplots-geom, echo=FALSE}
heights %>% ggplot(aes(sex, height)) +
  geom_boxplot()
```

## QQ-plots 

For qq-plots we use the `geom_qq` geometry. From the help file, we learn that we need to specify the `sample` (we will learn about samples in a later chapter). Here is the qqplot for men heights.

```{r ggplot-qq}
heights %>% filter(sex=="Male") %>%
  ggplot(aes(sample = height)) +
  geom_qq()
```

By default, the sample variable is compared to a normal distribution with average 0 and standard deviation 1. To change this,  from the help file, we use the `dparams` arguments. Adding an identity line is as simple as assigning another layer. For straight lines, we use the `geom_abline` function. The default is the identity.

```{r  ggplot-qq-dparams}
params <- heights %>% filter(sex=="Male") %>%
  summarize(mean = mean(height), sd = sd(height))

heights %>% filter(sex=="Male") %>%
  ggplot(aes(sample = height)) +
  geom_qq(dparams = params) +
  geom_abline()
```

Another option here is to scale the data first and then make a qqplot against the standard normal. 

```{r ggplot-qq-standard-units, eval=FALSE}
heights %>% 
  filter(sex=="Male") %>%
  ggplot(aes(sample = scale(height))) + 
  geom_qq() +
  geom_abline()
```

## Images

The two geometries used to create images are __geom_tile__ and __geom_raster__. They behave similarly, to see how they differ please consult the help file. To create an image in __ggplot2__ we need a data frame with the x and y coordinates as well as the values associated with each of these. Here is an a data frame.

```{r}
x <- expand.grid(x = 1:12, y = 1:10) %>% 
  mutate(z = 1:120) 
```

Note that this is the tidy version of a matrix `matrix(1:120, 12, 10)`. To plot the image we use the following code:

```{r, eval=FALSE}
x %>% ggplot(aes(x, y, fill = z)) + 
  geom_raster()
```

With these images you will often want to change the color scale. This can be done through the `scale_fill_gradientn` layer.

```{r ggplot2-image-new-colors}
x %>% ggplot(aes(x, y, fill = z)) + 
  geom_raster() + 
  scale_fill_gradientn(colors =  terrain.colors(10))
```





## Quick plots

In Section \@ref(qplot) we introduced `qplot` as a useful function  when we need to make a quick scatterplot. We can also use `qplot` to make histograms, density plots, boxplot, qqplots and more. Although it does not provide the level of control as `ggplot`, `qplot` is definitely useful as it permits us to make a plot with  short snipet of code. 

Suppose we have the female heights in an object `x`

```{r}
x <- heights %>% 
  filter(sex=="Male") %>% 
  pull(height)
```

To make a quick histogram we can use:

```{r qplot-example-1}
qplot(x)
```

The function guesses that we want to make a histogram because we only supplied one variable. In Section \@ref(qplot) we saw that if we supply `qplot` two variables it automatically makes a scatterplot. 

To make a quick qqplot you have to use the `sample` argument. Note that we can add layers just as we do with `ggplot`.

```{r qplot-example-2}
qplot(sample = scale(x)) + geom_abline()
```


If we supply a factor and a numeric vector we obtain a plot like the below. Note that in the code below we are using the `data` argument. Becuase the data frame is not the first argument in `qplot`, we have to use the dot operator.

```{r qplot-example-3}
heights %>% qplot(sex, height, data = .)
```

We can also select a specific geometry by using the `geom` argument. So to conver the plot above to a boxplot we use the following code:

```{r qplot-example-4}
heights %>% qplot(sex, height, data = ., geom = "boxplot")
```

We can also use the `geom` argument to generate a density plot instead of a histogram:

```{r qplot-example-5}
qplot(x, geom = "density")
```

Altough not as much as with `ggplot`, we do have some flexibility to improve the results of `qplot`. Looking at the help file we see several ways in which we can improve the look of the histogram above. Here is an example:

```{r qplot-example-6}
qplot(x, bins=15, color = I("black"), xlab = "Population")
```

**Technical note**: The reason we use `I("black")` is because we want `qplot` to treat `"black"` as a character rather than convert it to a factor, which is the default behavior within `aes`, which is internally called here. In general, the function `I` is used in R to say "keep it as it is".




### Exercises 


1. Now we are going to use the `geom_histogram` function to make a histogram of the heights in the `height` data frame. When reading the documentation for this function we see that it requires just one mapping, the values to be used for the histogram. Make a histogram of all the plots. 

    What is the variable containing the heights?

    A. `sex`
  
    B. `heights`
    
    C. `height`

    D. `heights$height`

2. Now create a ggplot object using the pipe to assign the heights data to a ggplot object an assign `height` to the x values through the `aes` function

3. Now we are ready to add a layer to actually make the histogram. Use the object created in the previous exercise and the `geom_histogram` function to make the histogram.

4. Note that when we run the code in the previous exercise we get the warning:
`stat_bin()` using `bins = 30`. Pick better value with `binwidth`.`

    Use the `binwidth` argument to change the histogram made in the previous exercise to use bins of size 1 inch. 

5. Now instead of a histogram we are going to make a smooth density plot. In this case we will not make an object, instead we will render the plot with one line of code. Change the code previously used

    ```{r, eval=FALSE}
    heights %>% 
      ggplot(aes(height)) +
      geom_histogram()
    ```

    to make a smooth density instead of a histogram.


6. Now we are going to make a density plot for males and females separately. We can do this using any of the `group` argument. We assign groups via the aesthetic mapping as each point needs to a group before making the calculations needed to estimate a density.

7. We can also assign groups through the `color` argument. This has the added benefit that it uses color to distinguish the groups. Change the code above to use color

8. We can also assign groups through the `fill` argument. This has the added benefit that it uses colors to distinguish the groups, like this:


    ```{r, eval=FALSE}
    heights %>% 
      ggplot(aes(height, fill = sex)) + 
      geom_density() 
    ```

    However, here the second density is drawn over the other. We can change this by using alpha blending. Set the alpha parameter to 0.2 in the `geom_density` function to make this change.












