## Case study: The Big Short

### Interest rates explained with chance model

More complex versions of the sampling models we have discussed are also used by banks to decide interest rates. Suppose you run a small bank that has a history of identifying potential homeowners that can be trusted to make payments. In fact, historically, in a given year, only 2% of your customers default, meaning that they don't pay back the money that you lent them. However, you are aware that if you simply loan money to everybody without interest, you will end up losing money due to this 2%. Although you know 2% of your clients will probably default, you don't know which ones. Yet by charging everybody just a bit extra in interest, you can make up the losses incurred due to that 2% and also cover your operating costs. You can also make a profit, but if you set the interest rates too high, your clients will go to another bank. We use all these facts and some probability theory to decide what interest rate you should charge.  

Suppose your bank will give out 1,000 loans for \$180,000 this year. Also, after adding up all costs, suppose your bank loses \$200,000 per foreclosure. For simplicity, we assume this includes all operational costs. A sampling model for this scenario can be coded like this:

```{r}
n <- 1000
loss_per_foreclosure <- -200000
p <- 0.02 
defaults <- sample( c(0,1), n, prob=c(1-p, p), replace = TRUE)
sum(defaults * loss_per_foreclosure)
```

Note that the total loss defined by the final sum is a random variable. Every time you run the above code, you get a different answer.  We can easily construct a Monte Carlo simulation to get an idea of the distribution of this random variable.

```{r}
B <- 10000
losses <- replicate(B, {
    defaults <- sample( c(0,1), n, prob=c(1-p, p), replace = TRUE) 
  sum(defaults * loss_per_foreclosure)
})
```

<!--
Here is the distribution of this random variable:
```{r losses-distribution, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
data.frame(losses_in_millions = losses/10^6) %>% ggplot(aes(losses_in_millions)) + geom_histogram(binwidth = 0.6, col="black")
```
-->

We don't really need a Monte Carlo simulation though. Using what we have learned, the CLT tells us that because our losses are a sum of independent draws, its distribution is approximately normal with expected value and standard errors given by:
 
```{r}
n*(p*loss_per_foreclosure + (1-p)*0)
sqrt(n)*abs(loss_per_foreclosure)*sqrt(p*(1-p))
```

We can now set an interest rate to guarantee that, on average, we break even. Basically, we need to add a quantity $x$ to each loan, which in this case are represented by draws, so that the expected value is 0. If we define $l$ to be the loss per foreclosure, we need:

$$
lp  + x(1-p) = 0
$$

which implies $x$ is 

```{r}
- loss_per_foreclosure*p/(1-p)
```

or an interest rate of `r - p*loss_per_foreclosure/(1-p) / 180000`.

However, we still have a problem. Although this interest rate guarantees that on average we break even, there is a 50\% chance that we lose money. If our bank loses money, we have to close it down. We therefore need to pick an interest rate that makes it unlikely for this to happen. At the same time, if the interest rate is too high, our clients will go to another bank so we must be willing to take some risks. So let's say that we want our chances of losing money to be 1 in 100, what does the $x$ quantity need to be now? This one is a bit harder. We want the sum $S$ to have:

$$\mbox{Pr}(S<0) = 0.01$$

We know that $S$ is approximately normal. The expected value of $S$ is 

$$\mbox{E}[S] = \{ lp + x(1-p)\}n$$ 

with $n$ the number of draws, which in this case represents loans. The standard error is 

$$\mbox{SD}[S] = |x-l| \sqrt{np(1-p)}.$$ 

Because $x$ is positive and $l$ negative $|x-l|=x-l$. Note that these are just an application of the formulas shown earlier, but using more compact symbols. 

Now we are going to use a mathematical "trick" that is very common in statistics. We add and subtract the same quantities to both sides of the event $S<0$ so that the probability does not change and we end up with a standard normal random variable on the left, which will then permit us to write down an equation with only $x$ as an unknown. This "trick" is as follows:

If $\mbox{Pr}(S<0) = 0.01$ then 
$$
\mbox{Pr}\left(\frac{S - \mbox{E}[S]}{\mbox{SE}[S]} < \frac{ - \mbox{E}[S]}{\mbox{SE}[S]}\right)
$$
And remember $\mbox{E}[S]$ and $\mbox{SE}[S]$ are the expected value and standard error of $S$, respectively. All we did above was add and divide by the same quantity on both sides.  We did this because now the term on the left is a standard normal random variable, which we will rename $Z$. 
Now we fill in the blanks with the actual formula for expected value and standard error:

$$
\mbox{Pr}\left(Z <  \frac{- \{ lp + x(1-p)\}n}{(x-l) \sqrt{np(1-p)}}\right) = 0.01
$$

It may look complicated, but remember that $l$, $p$ and $n$ are all known amounts, so eventually we will replace them with numbers. 

Now because the Z is a normal random with expected value 0 and standard error 1, it means that the quantity on the right side of the < sign must be equal to:

```{r}
qnorm(0.01)
```

for the equation to hold true. Remember that $z=$`qnorm(0.01)` gives us the value of $z$ for which: 

$$
\mbox{Pr}(Z \leq z) = 0.01
$$

So this means that the right side of the complicated equation must be $z$=`qnorm(0.01)`.

$$
\frac{- \{ lp + x(1-p)\}n} {(x-l) \sqrt{n p (1-p)}} = z
$$

The trick works because we end up with an expression containing $x$ that we know has to be equal to a known quantity $z$. Solving for $x$ is now simply algebra:

$$ x = - l \frac{ np  - z \sqrt{np(1-p)}}{n(1-p) + z \sqrt{np(1-p)}}$$

which is:
```{r}
l <- loss_per_foreclosure
z <- qnorm(0.01)
x <- -l*( n*p - z*sqrt(n*p*(1-p)))/ ( n*(1-p) + z*sqrt(n*p*(1-p)))
x
```

Our interest rate now goes up to `r x/180000`. This is still a very competitive interest rate. By choosing this interest rate, we now have an expected profit per loan of:

```{r}
loss_per_foreclosure*p + x*(1-p)
```

which is a total expected profit of about:

```{r}
n*(loss_per_foreclosure*p + x*(1-p)) 
```

dollars!

We can run a Monte Carlo simulation to double check our theoretical approximations:

```{r, cache=TRUE}
B <- 100000
profit <- replicate(B, {
    draws <- sample( c(x, loss_per_foreclosure), n, 
                        prob=c(1-p, p), replace = TRUE) 
    sum(draws)
})
mean(profit)
mean(profit<0)
```

### The Big Short

One of your employees points out that since the bank is making `r prettyNum(round(loss_per_foreclosure*p + x*(1-p)), big.mark=",")` dollars per loan, the bank should give out more loans! Why just $n$? You explain that finding those $n$ clients was hard. You need a group that is predictable and that keeps the chances of defaults low. He then points out that even if the probability of default is higher, as long as our expected value is positive, you can minimize your chances of losses by increasing $n$ and relying on the law of large numbers.

He claims that even if the default rate is twice as high, say 4%, if we set the rate just a bit higher than this value:
```{r}
p <- 0.04
r <- (- loss_per_foreclosure*p/(1-p)) / 180000
r
```

we will profit. At 5%, we are guaranteed a positive expected value of: 

```{r}
r <- 0.05
x <- r*180000
loss_per_foreclosure*p + x * (1-p)
```

and can minimize our chances of losing money by simply increasing $n$ since:

$$
\mbox{Pr}(S < 0) = 
\mbox{Pr}\left(Z < - \frac{\mbox{E}[S]}{\mbox{SE}[S]}\right)
$$
with $Z$ a standard normal random variable as shown earlier. If we define $\mu$ and $\sigma$ to be the expected value and standard deviation of the urn, respectively (that is of a single loan), using the formulas above we have: $\mbox{E}[S]= n\mu$ and  $\mbox{SE}[S]= \sqrt{n}\sigma$. So if we define $z$=`qnorm(0.01)`, we have: 
$$
 - \frac{n\mu}{\sqrt{n}\sigma} = - \frac{\sqrt{n}\mu}{\sigma} = z
$$
which implies that if we let:

$$
n \geq z^2 \sigma^2 / \mu^2
$$
we are guaranteed to have a probability of less than 0.01. The implication is that, as long as $\mu$ is positive, we can find an $n$ that minimizes the probability of a loss. This is a form of the law of large numbers: when $n$ is large, our average earnings per loan converges to the expected earning $\mu$.


With $x$ fixed, now we can ask what $n$ do we need for the probability to be 0.01? In our example, if we give out:

```{r}
z <- qnorm(0.01)
n <- ceiling((z^2*(x-l)^2*p*(1-p))/(l*p + x*(1-p))^2)
n
```

loans, the probability of losing is about 0.01 and we are expected to earn a total of 

```{r}
n*(loss_per_foreclosure*p + x * (1-p))
```

dollars! We can confirm this with a Monte Carlo simulation:


```{r}
p <- 0.04
x <- 0.05*180000
profit <- replicate(B, {
    draws <- sample( c(x, loss_per_foreclosure), n, 
                        prob=c(1-p, p), replace = TRUE) 
    sum(draws)
})
mean(profit)
```

This seems like a no brainer. As a result, your colleague decides to leave your bank and start his own high-risk mortgage company. A few months later, your colleague's bank has gone bankrupt. A book is written and eventually a movie is made relating the mistake your friend, and many others, made. What happened? 

Your colleague's scheme was mainly based on this mathematical formula:
$$
    \mbox{SE}[(X_1+X_2+\dots+X_n) / n] = \sigma / \sqrt{n}    
$$

By making $n$ large, we minimize the standard error of our per-loan profit. However, for this rule to hold, the $X$s must be independent draws: one person defaulting must be independent of others defaulting. Note that in the case of averaging the **same** event over and over, an extreme example of events that are not independent, we get a standard error that is $\sqrt{n}$ times bigger:
$$
    \mbox{SE}[(X_1+X_1+\dots+X_1) / n] =  \mbox{SE}[n X_1  / n] = \sigma > \sigma / \sqrt{n} 
$$

To construct a more realistic simulation than the original one your colleague ran, let's assume there is a global event that affects everybody with high-risk mortgages and changes their probability. We will assume that with 50-50 chance, all the probabilities go up or down slightly to somewhere between 0.03 and 0.05. But it happens to everybody at once, not just one person. These draws are no longer independent.

```{r}
p <- 0.04
x <- 0.05*180000
profit <- replicate(B, {
    new_p <- 0.04 + sample(seq(-0.01, 0.01, length = 100), 1)
    draws <- sample( c(x, loss_per_foreclosure), n, 
                        prob=c(1-new_p, new_p), replace = TRUE) 
    sum(draws)
})
```

Note that our expected profit is still large:

```{r}
mean(profit)
```

However, the probability of the bank having negative earnings shoots up to:

```{r}
mean(profit<0)
```

Even scarier is that the probability of losing more than 10 million dollars is:

```{r}
mean(profit < -10000000)
```

To understand how this happens look at the distribution:

```{r profit-distribution}
data.frame(profit_in_millions=profit/10^6) %>% 
  ggplot(aes(profit_in_millions)) + 
  geom_histogram(color="black", binwidth = 5)
```

The theory completely breaks down and the random variable has much more variability than expected. The financial meltdown of 2007 was due, among other things, to financial "experts" assuming independence when there was none.

## Exercises

1\. Create a random variable $S$ with the earnings of your bank if you give out 10,000 loans, the default rate is 0.3, and you lose \$200,000 in each foreclosure. Hint: use the code we showed in the previous section, but change the parameters.

2\. Run a Monte Carlo simulation with 10,000 outcomes for $S$. Make a histogram of the results. 


3\. What is the expected value of $S$?


4\. What is the standard error of $S$?

5\. Suppose we give out loans for \$180,000. What should the interest rate be so that our expected value is 0? 

6\. (Harder) What should the interest rate be so that the chance of losing money is 1 in 20? In math notation, what should the interest rate be so that $\mbox{Pr}(S<0) = 0.05$ ?

7\. If the bank wants to minimize the probabilities of losing money, which of the following does **not** make interest rates go up?  
    
a. A smaller pool of loans.
b. A larger probability of default.
c. A smaller required probability of losing money.
d. The number of Monte Carlo simulations.
