# Tidy data

```{r, message=FALSE}
library(dslabs)
ds_theme_set()
```

To help define tidy data, we refer to an example from the data visualization chapter in which we plotted fertility data across time for two countries: South Korea and Germany. To make the plot, we used this subset of the data:

```{r}
data("gapminder")
tidy_data <- gapminder %>% 
  filter(country %in% c("South Korea", "Germany")) %>%
  select(country, year, fertility)
head(tidy_data)
```

With the data in this format, we could quickly make the desired plot:

```{r tidy-example}
tidy_data %>% 
  ggplot(aes(year, fertility, color = country)) +
  geom_point()
```

One reason this code works seamlessly is because the data is _tidy_: each point is represented in a row. This brings us to the definition of _tidy data_: each row represents one observation and the columns represent the different variables available for each of these observations.

If we go back to the original data provided by GapMinder, we see that it does not start out _tidy_. We include an example file with the data shown in this graph mimicking the way it was originally saved in a spreadsheet:

```{r}
path <- system.file("extdata", package="dslabs")
filename <- file.path(path,  "fertility-two-countries-example.csv")
wide_data <- read_csv(filename)
```

The object `wide_data` includes the same information as the object `tidy_data` except it's in a different format: a `wide` format. Here are the first nine columns:

```{r}
select(wide_data, country, `1960`:`1967`)
```

Two important differences between the wide and tidy formats are: 1) in the wide format, each row includes several observations and 2) one of the variables, year, is stored in the header.
 
The ggplot code we introduced earlier no longer works here. For one, there is no `year` variable. To use the `tidyverse`, we need to wrangle this data into `tidy` format. The next two chapters describe tools that are available for _tidying_ data.


