# Tidy data

```{r, message=FALSE}
library(dslabs)
ds_theme_set()
```

To help define tidy data, we refer to an example from the data visualization chapter in which we plotted fertility data across time for two countries: South Korea and Germany. To make the plot, we used this subset of the data:

```{r}
data("gapminder")
tidy_data <- gapminder %>% 
  filter(country %in% c("South Korea", "Germany")) %>%
  select(country, year, fertility)
head(tidy_data)
```

With the data in this format, we could quickly make the desired plot:

```{r tidy-example}
tidy_data %>% 
  ggplot(aes(year, fertility, color = country)) +
  geom_point()
```

One reason this code works seamlessly is because the data is _tidy_: each point is represented in a row. This brings us to the definition of _tidy data_: each row represents one observation and the columns represent the different variables available for each of these observations.

If we go back to the original data provided by Gapminder, we see that it does not start out _tidy_. We include an example file with the data shown in this graph mimicking the way it was originally saved in a spreadsheet:

```{r}
path <- system.file("extdata", package="dslabs")
filename <- file.path(path,  "fertility-two-countries-example.csv")
wide_data <- read_csv(filename)
```

The object `wide_data` includes the same information as the object `tidy_data` except it's in a different format: a `wide` format. Here are the first nine columns:

```{r}
select(wide_data, country, `1960`:`1967`)
```

Two important differences between the wide and tidy formats are: 1) in the wide format, each row includes several observations and 2) one of the variables, year, is stored in the header.
 
The ggplot code we introduced earlier no longer works here. For one, there is no `year` variable. To use the __tidyverse__, we need to wrangle this data into `tidy` format. The next two chapters describe tools that are available for _tidying_ data.

## Exercises {-}


1. Examine the dataset `co2`. Which of the following is true:

    A. `co2` is tidy data: it has one year for each row.

    B. `co2` is not tidy: we need at least one column with a character vector.

    C. `co2` is not tidy:  it is a matrix not a data frame.

    D. `co2` is not tidy: to be tidy we would have to wrangle it to have three columns: year, month and value; then each co2 observation has a row.


2. Examine the dataset  `ChickWeight`. Which of the following is true:

    A.  `ChickWeight` is not tidy: each chick has more than one row. 

    B.  `ChickWeight` is tidy: each observation, here a weight, is represented by one row. The chick from which this measurement came from is one the variables.

    C.  `ChickWeight` is not a tidy: we are missing the year column.

    D.  `ChickWeight` is tidy: it is stored in a data frame.


3. Examine the dataset `BOD`. Which of the following is true:

    A. `BOD` is not tidy: it only has six rows.

    B. `BOD` is not tidy: the first column is just an index.

    C. `BOD` is tidy: each row is an observation with two values, time and demand.

    D. `BOD` is tidy: all small datasets are tidy by definition.


4. Which of the following datasets is tidy (you can pick more than one):

    A. `BJsales`

    B. `EuStockMarkets`

    C. `DNase`

    D. `Formaldehyde`

    E. `Orange`

    F. `UCBAdmissions`


