```{r, include=FALSE}
if(knitr::is_html_output()){
  knitr::opts_chunk$set(out.width = "500px", 
                        out.extra='style="display: block; margin-left: auto; margin-right: auto; background-color: #000; padding:3px;"')
} else{
  knitr::opts_chunk$set(out.width = "8in")
}
```

# GitHub 

GitHub is an online service that permits you to organize and share your code in what are called repositories using Git, a version control tool. In a later section, we will show you how to use Git and GitHub to organize your data science projects through RStudio. Before we do that, we install Git and then create an account and a repository. 

## Installing Git

In this Section we demonstrate how to install Git, the version control system we use to keep track of changes made to our code and to sync local copies of our code with copies hosted on GitHub. We also demonstrate how to install
Git Bash, a piece of software that emulates Unix on Windows machines. Note that Mac or Linux users should not install Git Bash.


```{r, echo=FALSE}
img_path <- "productivity/img/"
screenshots <- list.files(file.path(img_path, "windows-screenshots"))
screenshots <- file.path(img_path, "windows-screenshots", screenshots)
mac_screenshots <- list.files(file.path(img_path, "mac-screenshots"))
mac_screenshots <- file.path(img_path,"mac-screenshots", mac_screenshots)
```


The installation process for _Git_ is quite different for Mac and Windows. We include both below. 

Git is what we refer to as a _version control system_. These are useful for tracking changes to files as well as coordinating the editing of code by multiple collaborators. We will later learn how to use GitHub, which is a hosting system for code. You need Git to interact with GitHub. Having your code and, more generally, data science projects on GitHub is, among other things, a good way to showcase your work to current and potential employers. 

Git is most effectively used with Unix, although one can also use it through RStudio. In the next section, we describe Unix in more detail. Here we show you how to install software that permits you to use Git and Unix. The installation process is quite different for Windows and Mac, so we include two different sections.

### Installing Git and Git Bash on Windows

**Warning: The instructions in this subsection are not for Mac users.**

There are several pieces of software that will permit you to perform Unix commands on Windows. We will be using Git Bash as it interfaces with RStudio and it is automatically installed when we install Git for Windows.

1. Start by searching for _Git for windows_ on your browser and clicking on the link from _git-scm.com_.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[57]) 
```

2. This will take you to the _Download Git_ page from which you can download the _most recent maintained build_:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[58]) 
```

3. You can then accept to run the installer and agree to the license:

```{r, echo=FALSE, out.width = "4in",  out.extra=""}
knitr::include_graphics(screenshots[59:60]) 
```

4. In one of the installation steps, you will be asked to pick the default editor for Git. Unless you are already a _vi_ or _vim_ user, we recommend against selecting _vim_ which might be the default. If you do not recognize an editor you are familiar with among the options given, we recommend that you select _nano_ as your default editor for Git since it is the easiest to learn:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[61]) 
```

5. The next installation decision is actually an **important one**. This installation process installs Git Bash. We recommend that you select _Git and optional Unix tools from the Windows Command Prompt_ as this will permit you to learn Unix from within RStudio. However, if you do this, **some commands that run on your Windows command line will stop working**. If you do not use your windows command line, then this should not be a problem. Also, most, if not all, of these Windows command lines have a Unix equivalent that you will be able to use now. 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[62]) 
```

6. You can now continue selecting the default options.

    ```{r, echo=FALSE, out.width = "2.75in",  out.extra=""}
knitr::include_graphics(screenshots[63:65]) 
```

    ```{r, echo=FALSE, out.width = "2.75in",  out.extra=""}
knitr::include_graphics(screenshots[66:67]) 
```

    You have now installed Git on Windows.

7. A final and important step is to change a preference in RStudio so that Git Bash becomes the default Unix shell in RStudio. In RStudio, go to preferences (under the File pull down menu), then select _Terminal_, then select _Git Bash_:
 
```{r, echo=FALSE}
knitr::include_graphics(screenshots[68]) 
```

8. To check that you in fact are using Git Bash in RStudio, you can open a _New Terminal_ in RStudio: 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[69]) 
```

9. It should look something like this:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[70]) 
```


### Installing Git on the Mac

Before we show you the installation process, we introduce you to the _Terminal_. Macs already come with this terminal and it can be used to learn Unix. We can also use it to check if Git is already installed and, if not, start the installation process. 

1. To open a terminal, you can use command+spacebar and type terminal or you can find it in Utilities:

```{r, echo=FALSE, out.width = "2.75in",  out.extra=""}
knitr::include_graphics(mac_screenshots[1:2]) 
```
    
2. Once you start the terminal, you will see a console like this:

```{r, echo=FALSE}
knitr::include_graphics(mac_screenshots[3]) 
```

3. You might have Git installed already. One way to check is by asking for the version by typing:

    ```{bash, eval=FALSE, echo=TRUE}
git --version
```
    
    If you get a version number back, it is already installed. If not, you will get the following message:
    
    ```{r, echo=FALSE}
knitr::include_graphics(mac_screenshots[4]) 
```

    and you will be asked if you want to install it. You should click _Install_:

```{r, echo=FALSE}
knitr::include_graphics(mac_screenshots[5]) 
```

4. This will take you through the installation process:

```{r, echo=FALSE, out.width = "2.75in",  out.extra=""}
knitr::include_graphics(mac_screenshots[6:8]) 
```

5. Once installed, you can check for the version again and it should show you something like this:

```{r, echo=FALSE}
knitr::include_graphics(mac_screenshots[9]) 
```

Congratulations. You have installed Git on your Mac.

**Reminder**: On Windows, we install Git Bash. We do not need to do this on the Mac since we can use the pre-installed terminal to run Unix commands.

## GitHub Accounts

The next step is to get a GitHub account. Basic GitHub accounts are free. The first step is to go to [GitHub](github.com) where you will see a box in which you can sign up. 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[71]) 
```

You want to pick a name carefully. It should be short, easy to remember and to spell, somehow related to your name, and professional. This last one is important since you might be sending potential employers a link to your GitHub account. In the example below, I am sacrificing on the ease of spelling to incorporate my name. Your initials and last name are usually a good choice. If you have a very common name, then this may have to be taken into account. A simple solution would be to add numbers or spell out part of your name. 

The account I use for my research, _rafalab_, is the same one I use for my [webpage](http://rafalab.org) and [Twitter](http://twitter.com/rafalab), which makes it easy to remember for those that follow my work. 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[72]) 
```

Once you have a GitHub account, you are ready to connect RStudio to this account. You start by going to the _Global Options_ and selecting _Git/SVN_:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[73]) 
```

You then need to enter a path for the Git executable we just installed.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[74]) 
```

On the Windows default installation, this will be _C:/Program File/Git/bin/git.exe_, but you should find it by browsing your system as this can change from system to system. Now to avoid entering our GitHub password every time we try to access our repository, we will create what is called an _SSH RSA Key_. RStudio can do this for us automatically if we click on the _Create RSA Key_ button:


```{r, echo=FALSE}
knitr::include_graphics(screenshots[75]) 
```


```{r, echo=FALSE}
knitr::include_graphics(screenshots[76]) 
```

You can follow the default instructions as shown below: 

```{r, echo=FALSE, out.width = "3.5in",  out.extra=""}
knitr::include_graphics(screenshots[77:78]) 
```

RStudio and GitHub should now be able to connect and we are ready to create a first GitHub code repository.

## GitHub repositories

You are now ready to create a GitHub repository (repo). The general idea is that you will have at least two copies of your code: one on your computer and one on GitHub. If you add collaborators to this project, then each will have a copy on their computer. The GitHub copy is usually considered the _master_ copy that each collaborator syncs to. Git will help you keep all the different copies synced.

As mentioned, one of the advantages of keeping code on a GitHub repository is that you can easily share it with potential employers interested in seeing examples of your work. Because many data science companies use version control systems, like Git, to collaborate on projects, they might also be impressed that you already know at least the basics.

The first step in creating a repo for your code is to initialize on GitHub. Because you already created an account, you will have a page on GitHub with the URL _http://github.com/username_.

To create a repo, first log in to your account by clicking the _Sign In_ button on [https://github.com](http://github.com). You might already be signed in, in which case the _Sign In_ button will not show up:


```{r, echo=FALSE}
knitr::include_graphics(screenshots[79]) 
```

If signing in, you will have to enter your username and password. We recommend you set up your browser to remember this to avoid typing it in each time.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[80]) 
```

Once on your account, you can click on _Repositories_:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[81]) 
```

Then you click on _New_ to create a new repo:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[82]) 
```

You will then want to choose a good descriptive name for the project. In the future, you might have dozens of repos so keep that in mind when choosing a name. Here we will use _homerwork-0_. We recommend you make the repo public. If you want to keep it private, you will have to pay a monthly charge.


```{r, echo=FALSE}
knitr::include_graphics(screenshots[83]) 
```

You now have your first repo on GitHub. The next step will be to _clone it_ on your computer and start editing and syncing using Git. 

To do this, it is convenient to copy the link provided by GitHub specifically to connect to this repo, using Git as shown below. We will later need to copy and paste this so make sure to remember this step.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[84]) 
```


```{r, include=FALSE}
knitr::opts_chunk$set(out.width = NULL, out.extra = NULL)
```




