# Reproducible reports with R Markdown

The final product of a data analysis report is often a report. Many scientific publications can be thought of us a final report of a data analysis. The same is true for news articles based on data, an analysis report for your company, or lecture notes for a class on how to analyze data. The reports are often on paper or a PDF that includes a textual description of the finding along with some figures and tables resulting from the analysis. Imagine after you finish the analysis and the report, you are told that you were given the wrong dataset, You are sent you a new one and you are asked run the same analysis with this new dataset. Or what if you realize that a mistake was made and need to re-examine the code, fix the error and re-run the analysis? Or imagine that someone you are training wants to see the code and be able to reproduce the results to learn about the approach? Situations like the ones just described are actually quite common for a data scientist. Here, we describe how to generate reproducible reports with R markdown and knitR in a way that will greatly help with situations such as the ones described here. The main feature is that code and textual descriptions can be combined into the same document, and the figures and tables produced by the code are automatically added to the document. 



```{r, echo=FALSE}
screenshots <- list.files("productivity/img/windows-screenshots")
screenshots <- file.path("productivity/img/windows-screenshots", screenshots)
mac_screenshots <- list.files("productivity/img/mac-screenshots")
mac_screenshots <- file.path("productivity/img/mac-screenshots", mac_screenshots)
```


## R Markdown
R markdown is a format for _literate programming_ document. It is based on _markdown_, a markup language that is widely used to generate html pages. You can learn more about markdown [here](https://www.markdowntutorial.com/). Literate programming weaves instructions, documentation and detailed comments in between machine executable code, producing a document that describes the program that is best for human understanding (Knuth 1984). Unlike a word processor, such as Microsoft Word, where you what you see is what you get, with R markdown you need to _compile_ the document into the final report. The R markdown document looks different than the final product. This seems like a disadvantage at first, but it is not at all because, for example, instead of producing plots and inserting them one by one into the word processing document, the plots are automatically added. 

In RStudio, you can start an R Markdown document by clicking on _File_, _New File_, the _R Markdown_, like this:

```{r, echo=FALSE,  out.width = "500px", out.extra='style="display: block; margin-left: auto; margin-right: auto; background-color: #000; padding:3px;"'}
knitr::include_graphics(screenshots[104]) 
```

You will then be asked to enter a title and author for your document. We are going to prepare a report on gun murders so we will give it an appropriate name. You can also decide what format you would like the final report to be in: HTML, PDF, or Microsoft Word. Later we can easily change this, but here we select html as it is the preferred format for debugging purposes:

```{r, echo=FALSE,  out.width = "500px", out.extra='style="display: block; margin-left: auto; margin-right: auto; background-color: #000; padding:3px;"'}
knitr::include_graphics(screenshots[105]) 
```

This will generate a template file:

```{r, echo=FALSE,  out.width = "500px", out.extra='style="display: block; margin-left: auto; margin-right: auto; background-color: #000; padding:3px;"'}
knitr::include_graphics(screenshots[106]) 
```

As a convention, we use the `Rmd` suffix for these files.

Once you gain experience with R Markdown, you will be able to do this without the template and can simply start from a blank template.

In the template, you will see several things to note.



### The Header

At the top you see:

```
---
title: "Report on Gun Murders"
author: "Rafael Irizarry"
date: "April 16, 2018"
output: html_document
---
```

The things between the `---` is the header. We actually don't need a header, but it is often useful. You can define many other things in the header than what is included in the template. We don't discuss those here, but much information is available online. The one parameter that we will highlight is `output`. By changing this to, say, `pdf_document`, we can control the output that is produced when we compile. 

### R code chunks

In various places in the document, we see something like this:

<pre><code>```{r}
summary(pressure)
```
</code></pre>

To add your own R chunks, you can type the characters above quickly with the key binding command-option-I on the Mac and Ctrl-Alt-I on Windows.


These are the code chunks. When you compile the document, the R code inside the chunk, in this case `summary(pressure)`, will be evaluated and the result included in that position in the final document. 

This applies to plots as well; the plot will be placed in that position. We can write something like this:

<pre><code>```{r} 
plot(pressure)
```
</pre></code>


By default, the code will show up as well. To avoid having the code show up, you can use an argument. To avoid this, you can use the argument `echo=FALSE`. For example:

<pre><code>```{r, echo=FALSE} 
plot(pressure)
```
</pre></code>

We recommend getting into the habit of adding a label to the R code chunks. This will be very useful when debugging, among other situations. You do this by adding a descriptive word like this:

<pre><code>```{r pressure-summary} 
summary(pressure)
```
</pre></code>

### Global options

One of the R chunks contains a complex looking call:


<pre><code>```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
</pre></code>

We will not cover this here, but as you become more experienced with R Markdown, you will learn the advantages of setting global options for the compilation process.

```## added for editing purposes


## knitR

We use the `knitr` package to compile R. The specific function used to compile is `knit` function takes a filename as input. [AN: review previous sentence] RStudio provides a button that actually makes it easier to compile the document. For the screenshot below, we have edited the document so that a report on gun murder is produced. You can see the file [here](https://raw.githubusercontent.com/rairizarry/murders/master/report.Rmd). You can now click on the `Knit` button:

```{r, echo=FALSE,  out.width = "500px", out.extra='style="display: block; margin-left: auto; margin-right: auto; background-color: #000; padding:3px;"'}
knitr::include_graphics(screenshots[107]) 
```


The first time you click on the _Knit_ button, a dialog box may appear asking you to install packages you need:

```{r, echo=FALSE,  out.width = "500px", out.extra='style="display: block; margin-left: auto; margin-right: auto; background-color: #000; padding:3px;"'}
knitr::include_graphics(screenshots[108]) 
```


Once you have installed the packages, clicking the _Knit_ will compile your R markdown file and the resulting document will pop-up:

```{r, echo=FALSE,  out.width = "500px", out.extra='style="display: block; margin-left: auto; margin-right: auto; background-color: #000; padding:3px;"'}
knitr::include_graphics(screenshots[109]) 
```

This produces an html document which you can see in your working directory. To view it, open a terminal and list the files. You can open the file in a browser and use this to present your analysis. You can also produce a PDF or Microsoft document by changing: 

`output: html_document` to `output: pdf_document` or `output: word_document`. 

We can also produce documents that render on GitHub using `output: github_document` like this:


```{r, echo=FALSE,  out.width = "500px", out.extra='style="display: block; margin-left: auto; margin-right: auto; background-color: #000; padding:3px;"'}
knitr::include_graphics(screenshots[110]) 
```

This will produce a markdown file, with suffix `md`, that renders in GitHub. Because we have uploaded these files to GitHub you can click on the `md` file:


```{r, echo=FALSE,  out.width = "500px", out.extra='style="display: block; margin-left: auto; margin-right: auto; background-color: #000; padding:3px;"'}
knitr::include_graphics(screenshots[124]) 
```

and you will see the report as a webpage:


```{r, echo=FALSE,  out.width = "500px", out.extra='style="display: block; margin-left: auto; margin-right: auto; background-color: #000; padding:3px;"'}
knitr::include_graphics(screenshots[125]) 
```

This is a convenient way to share your reports.


## More on R markdown

There is a lot more you can do with R markdown. We highly recommend you continue learning as you gain more experience writing reports in R. There are many free resources on the internet including:

* [RStudio's tutorial](https://rmarkdown.rstudio.com)
* [The cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
* [The knitR book](https://yihui.name/knitr/)


