```{r, include=FALSE}
if(knitr::is_html_output()){
  knitr::opts_chunk$set(out.width = "500px", 
                        out.extra='style="display: block; margin-left: auto; margin-right: auto; background-color: #000; padding:3px;"')
} else{
  knitr::opts_chunk$set(out.width = "8in")
}
```

# RStudio

```{r, echo=FALSE}
img_path <- "productivity/img/"
screenshots <- list.files(file.path(img_path, "windows-screenshots"))
screenshots <- file.path(img_path, "windows-screenshots", screenshots)
mac_screenshots <- list.files(file.path(img_path, "mac-screenshots"))
mac_screenshots <- file.path(img_path,"mac-screenshots", mac_screenshots)
```

RStudio will be our launching pad for data science projects. It not only provides an editor for us to create and edit our scripts but many other useful tools. In this section, we go over some of the basics.

## The panes

When you start RStudio for the first time, you will see three panes. The left pane shows the R console. On the right, the top pane includes three tabs: _Environment_, _History_, and _Connections_, while the bottom pane shows five tabs: _File_, _Plots_, _Packages_, _Help_ and _Viewer_. You can click on each tab to move across the different features. 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[30]) 
```

To start a new script, you can click on File, the New File, then R Script. 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[31]) 
```

This starts a new pane on the left and it is here where you can start writing your script. 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[32]) 
```

## Key bindings

Many tasks we perform with the mouse can be achieved with a combination of key strokes instead. These keyboard versions for performing tasks are referred to as _key bindings_. For example, we just showed how to use the mouse to start a new script, but you can also use a key binding: Ctrl+Shift+N on Windows and command+shift+N on the Mac. 

Although in this tutorial we often show how to use the mouse, **we highly recommend that you memorize key bindings for the operations you use most**. RStudio provides a useful cheat sheet with the most widely used commands. You can get it from RStudio directly:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[33]) 
```

and it looks like this:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[34]) 
```

You might want to keep this handy so you can look up key-bindings when you find yourself performing repetitive point-and-click tasks.

## Installing R packages

Most of what we have learned in this book depends on the tidyverse packages. The data we have been working on depend on the __dslabs__ package. These packages do not come pre-installed in R. In fact, the default installation of R is quite minimal and, for many of your projects, you will need to download and install one or more packages.

You can install packages directly from R with the command `install.packages()`. To install the tidyverse package, we would type, in the R console: 

```{r, eval=FALSE, echo=TRUE}
install.packages("tidyverse")
```

We can install more than one package at once by feeding a character vector to this function:

```{r, eval=FALSE, echo=TRUE}
install.packages(c("tidyverse", "dslabs"))
```

You can also install packages using RStudio in the following way: 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[35]) 
```

One advantage of using RStudio is that it auto-completes package names once you start typing, which is helpful when you do not remember the exact spelling of the package:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[36]) 
```

Once you select your package, we recommend selecting all the defaults:

```{r, echo=FALSE, out.width = "4in",  out.extra="" }
knitr::include_graphics(screenshots[37:38]) 
```

Remember that installing __tidyverse__ actually installs several packages.

Once packages are installed, you can load them into R and you do not need to install them again, unless you install a fresh version of R. Remember packages are installed in R not RStudio.

It is helpful to keep a list of all the packages you need for your work in a script because if you need to perform a fresh install of R, you can re-install all your packages by simply running a script.

You can see all the packages you have installed using the following function:

```{r, eval=FALSE, echo=TRUE}
installed.packages()
```


## Running commands while editing scripts

There are many editors specifically made for coding. These are useful because color and indentation are automatically added to make code more readable. RStudio is one of these editors, and it was specifically developed for R. One of the main advantages provided by RStudio over other editors is that we can test our code easily as we edit our scripts. Below we show an example.

Let's start by opening a new script as we did before. A next step is to give the script a name. We can do this through the editor by saving the current new unnamed script. To do this, click on the save icon or using the key binding Ctrl+S on Windows and command+S on the Mac.  

```{r, echo=FALSE}
knitr::include_graphics(screenshots[39]) 
```

When you ask for the document to be saved for the first time, RStudio will prompt you for a name. You want to use a descriptive name, with lower case letters, no spaces, only hyphens to separate words, and then followed by the suffix _.R_. We will call this script _my-first-script.R_.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[40]) 
```

Now we are ready to start editing our first script. The first lines of code in an R script are dedicated to loading the libraries we will use. Another useful RStudio feature is that once we type `library()` it starts auto-completing with libraries that we have installed. Note what happens when we type `library(ti)`:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[41]) 
```

Another feature you may have noticed is that when you type `library(` the second parenthesis is automatically added. This will help you avoid one of the most common errors in coding: forgetting to close a parenthesis.

Now we can continue to write code. As an example, we will make a graph showing murder totals versus population totals by state. Once you are done writing the code needed to make this plot, you can try it out by _sourcing_ in the code. To do this, click on the _Run_ button on the upper right side of the editing pane. You can also use the key binding: Ctrl+Shift+Enter on Windows or command+shift+return on the Mac.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[42]) 
```

Once you run the code below, you will see it appear in the R console and, in this case, the generated plot appears in the plots console. Note that the plot console has a useful interface that permits you to click back and forward across different plots, zoom in to the plot, or save the plots as files.

```r
library(tidyverse)
library(dslabs)
data(murders)

murders %>%
  ggplot(aes(population, total, label = abb, color = region)) +
  geom_label()
```

```{r, echo=FALSE}
knitr::include_graphics(screenshots[43]) 
```

To run one line at a time instead of the entire script, you can use Control+Enter or Windows and Command+Return on the Mac.

## Global options

You can change the look and functionality of RStudio quite a bit. 

To change the global options you click on _Tools_ then _Global Options..._. 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[44]) 
```

As an example, we show how to change the appearance of the editor. To do this click on _Appearance_ and then notice the _Editor theme_ options.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[45]) 
```

You can click on these and see examples of how your editor will look. 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[46]) 
```

I personally like the _Cobalt_ option. This makes your editor look like this:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[47]) 
```

As a second example, we show how to make a change that we **highly recommend**. This is to change the _Save workspace to .RData on exit_ to _Never_ and uncheck the _Restore .RData into workspace at start_. By default, when you exit R saves all the objects you have created into a file called .RData. This is done so that when you restart the session in the same folder, it will load these objects. We find that this causes confusion especially when we share code with colleagues and assume they have this .RData file. To change these options, make your _General_ settings look like this:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[48]) 
```

## Keeping organized with RStudio projects

A data analysis project is not always a dataset and a script. It often involves several scripts, the data may be saved across several files, and it is often convenient to save intermediate files. RStudio projects provide a way to keep all this organized in one folder. We will later learn how RStudio facilitates sharing work in these projects.

To organize yourself on a computer, it is essential that you understand your filesystem. A systematically organized filesystem can greatly increase your productivity, especially if you work on more than one project at at time. In a later section, we explain how Unix provides a powerful tool to help you with this. In this section, we will create a folder in a default location for illustrative purposes. Once you become a regular R user, you will want to think carefully about the best location for the folder in which you will keep a new project.

To start a project, click on _File_ and then _New Project_
```{r, echo=FALSE}
knitr::include_graphics(screenshots[49]) 
```

Unless you have a pre-selected folder to save the work, you will select the _New Directory_ option. 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[50]) 
```

Then, for a data analysis project, you usually select the  _New Project_ option:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[51]) 
```

Now you will have to decide on the location of the folder that will be associated with your project, as well as the name of the folder. When choosing a folder name, just like with file names, make sure it is a meaningful name that will help you remember what the project is about. As with files, we recommend using lower case letters, no spaces, and hyphens to separate words. We will call the folder for this project _my-first-project_. This will then generate a file called _my-first-project.Rproj_ in the folder associated with the project. We will see how this is useful a few lines below.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[52]) 
```

You will be given options on where this folder should be on your file system. In this example, we will place it in our home folder, but this is generally not good practice. As we describe in more detail later, you want to organize your file system following a hierarchical approach and you might have a folder called _projects_ where you keep a folder for each project.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[53]) 
```

When you start using RStudio with a project, you will see the project name in the upper left corner. This will remind you what project this particular RStudio session belongs to. When you open an RStudio session with no project, it will say _Project: (None)_. 

When working on a project, all files will be saved and searched for in the folder associated with the project. Below, we show an example of a script that we wrote and saved with the name _code.R_. Because we used a meaningful name for the project, we can be a bit less informative when we name the files. Although we do not do it here, you can have several scripts open at once. You simply need to click _File_, then _New File_ and pick the type of file you want to edit.


```{r, echo=FALSE}
knitr::include_graphics(screenshots[54]) 
```

One of the main advantages of using Projects is that after closing RStudio, if we wish to continue where we left off on the project, we simply double click or open the file saved when we first created the RStudio project. In this case, the file is called _my-first-project.Rproj_. 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[55]) 
```

If we open this file, RStudio will start up and open the scripts we were editing.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[56]) 
```


## Using Git and GitHub in RStudio

We are now ready to clone a repo, start editing files on our computer and syncing to GitHub. We will use RStudio to facilitate this. We will also use Unix for the first time! 

A first step is to let Git know who we are. This will make it easier to connect with GitHub. We start by opening a terminal window in RStudio (remember you can get one through _Tools_ in the menu bar). Now we use the `git config` command to tell Git who we are. We will type the following two commands in our terminal window:

```{bash, eval=FALSE, echo=TRUE}
git config --global user.name "Your Name"
git config --global user.mail "your@email.com"
```

You need to use the email account that you used to open your GitHub account. The RStudio sessions should look something like this:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[85]) 
```

Now we are ready to start a RStudio project that uses version control and stores the code on a GitHub repo. To do this, we start a project but, instead of _New Directory_, we will select _Version Control_:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[86]) 
```

Then we will select _Git_ as our version control system:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[87]) 
```

The repository URL is the link you used to clone. Above, we used `https://github.com/username/homework-0.git` as an example. In the project directory name, you need to put the name of the folder that was generated, which in our example will be the name of the repo `homework-0`. This will create a folder called `homework-0` on your local system.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[88]) 
```

Once you do this, the project is created and it is aware of the connection to a GitHub repo. You will see on the top right corner the name and type of project as well as a new tab on the upper right pane titled _Git_. 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[89]) 
```

If you select this tab, it will show you the files on your project with some icons that give you information about these files and their relationship to the repo. In the example below, we already added a file to the folder, called _code.R_ which you can see in the editing pane.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[90]) 
```

We now need to pay attention to the Git pane. It is important to know that **your local files and the GitHub repo will not be synched automatically**.  You have to sync when you are ready and we show you how below. To truly understand why RStudio does not automatically sync, we need to learn more details about Git, and we will do so in Section \@ref(git). Right now, we will quickly show you how to sync with this simple example. 

The main actions in Git are to: 

1. **pull** changes from the remote repo, in this case the GitHub repo,
2. **add** files, or as we say in the Git lingo _stage_ files, 
3. **commit** changes to the local repo and 
4. **push** changes to the _remote_ repo, in our case the GitHub repo.

Before we start working on a collaborative project, usually the first thing we do is _pull_ in the changes from the remote repo, in our case the one on GitHub. However, for the example shown here, since we are starting with an empty repo and we are the only ones making changes, we don't need to start by pulling.


In RStudio, the status of the file as it relates to the remote and local repos are represented in the status symbols with colors. A yellow square means that Git knows nothing about this file. To sync with the GitHub repo, we need to _add_ the file, then _commit_ the change to our local Git repo, then _push_ the change to the GitHub repo. Right now, the file is just on our computer. To add the file using RStudio, we click the _Stage_ box. You will see that the status icon now changes to a green A. 


```{r, echo=FALSE}
knitr::include_graphics(screenshots[91]) 
```

Note: we are only adding the _code.R_ file. We don't necessarily need to add all the files in our local repo to the GitHub repo; only the ones we want to keep track of or the ones we want to share. If our work is producing files of a certain type that we do not want to keep track of, we can add the suffix that defines these files to the .gitignore file. More details on using .gitignore are included [here](https://git-scm.com/docs/gitignore). These files will stop appearing in your RStudio Git pane. For the example shown here, we will only be adding the _code.R_. But, in general, for an RStudio project, we recommend adding both the .gitignore and .Rproj files.

Now we are ready to commit the file to our local repo. In RStudio, we can use the _Commit_ button. This will open a new dialog window. With Git, whenever we commit a change, we are required to enter a comment describing the changes being _committed_. 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[92]) 
```

In this case, we will simply describe that we are adding a new script. In this dialog box, RStudio also gives you a summary of what you are changing to the GitHub repo. In this case, because it is a new file, the entire file is highlighted as green, which highlights the changes.

Once we hit the commit button, we should see a message from Git with a summary of the changes that were committed.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[93]) 
```

Now we are ready to _push_ these changes to the GitHub repo. We can do this by clicking on the _Push_ button on the top right corner:


```{r, echo=FALSE}
knitr::include_graphics(screenshots[94]) 
```

We now see a message from Git letting us know that the push has succeeded.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[95]) 
```

In the pop-up window we no longer see the `code.R` file. This is because no new changes have been performed since we last pushed. We can exit this pop-up window now and continue working on our code.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[96]) 
```

If we now visit our repo on the web, we will see that it matches our local copy. 

```{r, echo=FALSE}
knitr::include_graphics(screenshots[97]) 
```

Congratulations you have successfully created a GitHub code repository! Soon we will learn how to share our code. But before we continue learning about Git, we will provide a brief introduction to Unix and how it is used to keep organized. 

## Running R without RStudio

Although we highly recommend that beginners use R through RStudio, you can use R without RStudio. 

You can start it like any other program. 

If you followed the default installation on Windows, a shortcut will appear on your desktop which you can click to start R.

```{r, echo=FALSE}
knitr::include_graphics(screenshots[16]) 
```

On the Mac, R will be in the Application folder. 

```{r, echo=FALSE, out.width = "2.75in",  out.extra=""}
knitr::include_graphics(mac_screenshots[31:32]) 
```

If you start R without RStudio, you will see an R console in which you can start typing commands:

```{r, echo=FALSE}
knitr::include_graphics(screenshots[17]) 
```

But we will be much more productive using an editor developed for coding, such as the one provided by RStudio. In the next section, we demonstrate how to install RStudio.


```{r, include=FALSE}
knitr::opts_chunk$set(out.width = NULL, out.extra = NULL)
```

