# Data types

Variables in R can be of different types. For example, we need to distinguish numbers from character strings and tables from simple lists of numbers. The function `class` helps us determine what type of object we have:

```{r}
a <- 2
class(a)
```

To work efficiently in R, it is important to learn the different types of variables and what we can do with these.

## Data frames

Up to now, the variables we have defined are just one number. This is not very useful for storing data. The most common way of storing a dataset in R is in a _data frame_. Conceptually, we can think of a data frame as a table with rows representing observations and the different variables reported for each observation defining the columns. Data frames are particularly useful for datasets because we can combine different data types into one object. 

We stored the data for our motivating example in a data frame. You can access this dataset by loading the __dslabs__ library and loading the `murders` dataset using the `data` function:

```{r}
library(dslabs)
data(murders)
```

To see that this is in fact a data frame, we type:

```{r}
class(murders)
```

## Examining an object

The function `str` is useful for finding out more about the structure of an object:

```{r}
str(murders)
```

This tells us much more about the object. We see that the table has 51 rows (50 states plus DC) and five variables. We can show the first six lines using the function `head`:

```{r}
head(murders)
```

In this dataset, each state is considered an observation and five variables are reported for each state.

Before we go any further in answering our original question about different states, let's learn more about the components of this object.

## The accessor: `$`

For our analysis we will need to access the different variables represented by columns included in this data frame. To do this, we use the accessor operator `$` in the following way:

```{r}
murders$population
```

But how did we know to use population? Previously, by applying the function `str` to the object `murders`, we revealed the names for each of the five variables stored in this table. We can quickly access the variable names using:

```{r}
names(murders)
```

It is important to know that the order of the entries in `murders$population` preserves the order of the rows in our data table. This will later permit us to manipulate one variable based on the results of another. For example, we will be able to order the state names by the number of murders.

**Tip**: R comes with a very nice auto-complete functionality that saves us the trouble of typing out all the names. Try typing `murders$p` then hitting the _tab_ key on your keyboard. This functionality and many other useful auto-complete features are available when working in RStudio.

## Vectors: numerics, characters, and logical

The object `murders$population` is not one number, but several. We call these types of objects _vectors_. A single number is technically a vector, but, in general, we use vectors to refer to objects with several entries. The function `length` tells you how many entries are in the vector:

```{r}
pop <- murders$population
length(pop)
```

This particular vector is _numeric_ since population sizes are numbers:

```{r}
class(pop)
```
In a numeric vector, every entry must be a number. 

To store character strings, vectors can also be of class _character_. For example, the state names are characters:

```{r}
class(murders$state)
```

As with numeric vectors, all entries in a character vector need to be a character.

Another important type of vectors are _logical vectors_. These must be either `TRUE` or `FALSE`. 

```{r}
z <- 3 == 2
z
class(z)
```

Here the `==` is a relational operator asking if 3 is equal to 2. Remember that in R, if you just use one `=`, you actually assign. 

You can see the other _relational operators_ by typing:

```{r, eval=FALSE}
?Comparison
```

In future sections, you will see how useful relational operators can be.


**Advanced**: Mathematically, the values in `pop` are integers and there is an integer class in R. However, by default, numbers are assigned class numeric even when they are round integers. For example, `class(1)` returns numeric. You can turn them into class integer with the `as.integer()` function or by adding an `L` like this: `1L`. Note the class by typing: `class(1L)`


## Factors

In the `murders` dataset, we might expect the region to also be a character vector. However, it is not:

```{r}
class(murders$region)
```

It is a _factor_. Factors are useful for storing categorical data. We can see that there are only 4 regions by using the `levels` function:


```{r}
levels(murders$region)
```

In the background, R stores these _levels_ as integers and keeps a map to keep track of the labels. This is more memory efficient than storing all the characters. Yet factors are also a source of confusion since sometimes they behave like characters and sometimes they do not. As a result, confusing factors and characters are a common source of bugs. 

In general, we recommend avoiding factors as much as possible, although they are sometimes necessary to fit models containing categorical data. We will see several example of these faction in the Regression and Machine Learning parts.
 
## Lists

Data frames are a special case of _lists_. We will cover lists in more detail later, but know that they are useful because you can store any combination of different types. Below is an example of a list we created for you:


```{r, echo=FALSE}
record <- list(name = "John Doe",
             student_id = 1234,
             grades = c(95, 82, 91, 97, 93),
             final_grade = "A")
```

```{r}
record
class(record)
```

We won't be using lists until later, but you might encounter one in your own exploration of R. For this reason, we show you some basics here. 

As with data frames, you can extract the components of a list with the accessor `$`. In fact, data frames are a type of list.

```{r}
record$student_id
```

We can also use double brackets like this:

```{r}
record[["student_id"]]
```

You should get used to the fact that in R, there are often several ways to do the same thing, such as accessing entries.

## Matrices  

Matrices are another type of object that are common in R. Matrices are similar to data frames in that they are two dimensional: they have rows and columns. However, like numeric, character and logical vectors, entries in matrices have to be all the same type. For this reason data frames are much more useful for storing data, since we can have characters, factors and numbers in them. 

Yet matrices have a major advantage over data frames: we can perform a matrix algebra operations, a powerful type of mathematical technique. We do not describe these operations in this book, but much of what happens in the background when you perform a data analysis involves matrices. We cover matrices in more detail in Chapter \@ref(matrix-algebra) but describe them briefly here since some of the functions we will learn return matrices. 

We can define a matrix using the `matrix` function. We need to specify the number of rows and columns.

```{r}
mat <- matrix(1:12, 4, 3)
mat
```

You can access specific entries in a matrix using `[`. If you want the second row, third column, you use:

```{r}
mat[2, 3]
```

If you want the entire second row, you leave the column spot empty:

```{r}
mat[2, ]
```

Notice that this returns a vector, not a matrix.

Similarly, if you want the entire third column, you leave the row spot empty:

```{r}
mat[, 3]
```

This is also a vector, not a matrix.

You can access more than one column or more than one row if you like. This will give you a new matrix.

```{r}
mat[, 2:3]
```

You can subset both rows and columns:

```{r}
mat[1:2, 2:3]
```

We can convert matrices into data frames using the function `as.data.frame`:

```{r}
as.data.frame(mat)
```

You can also use the `[` to access rows and columns of a data frame:

```{r}
data("murders")
murders[25, 1]
murders[2:3, ]
```


## Exercises 

1. Load the US murders dataset.

    ```{r}
    library(dslabs)
    data(murders)
    ```
    
    Use the function `str` to examine the structure of the `murders` object. We can see that this object is a data frame with 51 rows and five columns. Which of the following best describes the variables represented in this data frame?
    
    A. The 51 states.
    
    B. The murder rates for all 50 states and DC.
    
    C. The state name, the abbreviation of the state name, the state's region, and the state's population and total number of murders for 2010. 
    
    D. `str` shows no relevant information.

2. What are the column names used by the data frame for these five variables?
    
3. Use the accessor `$` to extract the state abbreviations and assign them to the object `a`. What is the class of this object?

4. Now use the square brackets to extract the state abbreviations and assign them to the object `b`. Use the `identical` function to determine if `a` and `b` are the same.

5. We saw that the `regions` column stores a factor. You can corroborate this by typing: 

    ```{r, eval=FALSE}
    class(murders$region)
    ```

    With one line of code, use the function `levels` and `length` to determine the number of regions defined by this dataset.

6. The function `table` takes a vector and returns the frequency of each element. You can quickly see how many states are in each region by applying this function. Use this function in one line of code to create a table of states per region.

    
