# Importing data

In this chapter, we used a data set already stored in an R object. A data scientist will rarely have such luck and will have to import data into R from either a file, a database, or other source. We cover this in more detail in the Data Wrangling part of the book. But because it is so common to read data from a file, we will briefly describe the key approach and function, in case you want to use your new knowledge on one of your own datasets.


Small datasets such as the one used in this chapter are commonly stored as Excel files. Although there
are R packages designed to read Excel (xls) format, you generally want
to avoid this format and save files as comma delimited (Comma-Separated
Value/CSV) or tab delimited (Tab-Separated Value/TSV/TXT) files.
These plain-text formats make it easier to share data since commercial software is not required for working with the data.


## Paths and the working directory 

The first step is to find the file containing your data and know its *path*.
When you are working in R, it is useful to know your _working directory_. This is the folder in which R will save or look for files by default. You can see your working directory by typing:

```{r, eval=FALSE}
getwd()
```

You can also change your working directory using the function `setwd` or you can change it through RStudio by clicking on "Session".

The functions that read and write files (there are several in R) assume you mean to look for files or write files in the working directory. Our recommended approach for beginners will have you reading and writing to the working directory. However, you can also type the [full path](http://www.computerhope.com/jargon/a/absopath.htm), which will work independently of the working directory.

We have included the US murders data in CSV file as part of the __dslabs__ package. We recommend placing your data in your working directory. Because knowing where packages store files is rather advanced, we provide the following code that finds the directory and copies the file:

```{r}
dir <- system.file(package="dslabs") #extracts the location of package
filename <- file.path(dir,"extdata/murders.csv")
file.copy(filename, "murders.csv")
```

If a file is copied succesfully, the `file.copy` function returns `TRUE`.

You should be able to see the file in your working directory and can check by using:

```{r, eval=FALSE}
list.files()
```

## `read.csv`

We are ready to read in the file. There are several functions for reading in tables. Here we introduce one included in base R:

```{r}
dat <- read.csv("murders.csv")
head(dat)
```

We can see that we have read in the file.

Warning: `read.csv` automatically converts characters to factors. Note, for example, that: 

```{r}
class(dat$state)
```

You can avoid this using:
```{r}
dat <- read.csv("murders.csv", stringsAsFactors = FALSE)
class(dat$state)
```

With this call, the region variable is no longer a factor, but we can easily change this with:

```{r}
dat <- mutate(dat, region =  as.factor(region))
```

Now that we are done with this example, we remove the `murders.csv` file from our working directory

```{r}
file.remove("murders.csv")
```

If the file is removed successfully `file.remove` returns `TRUE`.

