# Matrices

Matrices are another type of object that are common in R. Matrices are similar to data frames in that they are two dimensional: they have rows and columns. However, like numeric, character and logical vectors, entries in matrices have to be all the same type. For this reason data frames are much more useful for storing data, since we can have multiple data types in them. 

Yet matrices have a major advantage over data frames: we can perform a matrix algebra operations, a powerful type of mathematical technique. We do not describe these operations in this book, but much of what happens in the background when you perform a data analysis involves matrices. For this reason, and others, some of the functions we will learn return matrices. We will describe them briefly here.

One way matrices show up is if a function called with `sapply` returns two or more values. For example, suppose we write a function that for any number `x` returns `x`, `x^2` and `x^3`:

```{r}
cubic <- function(x) {
  c(x, x^2, x^3)
}
cubic(2)
```

If we use `sapply` to compute these three numbers for each of the elements in a vector, R automatically _binds_ the results as columns in a matrix:


```{r}
n <- seq(1,5)
mat <- sapply(n, cubic)
mat
```


## Accessing matrix values

You can access specific entries in a matrix using `[`. If you want the value from the second row, third column, you use:

```{r}
mat[2,3]
```

If you want the entire second row, leave the column spot empty:

```{r}
mat[2, ]
```

Notice that this returns a vector, not a matrix.

Similarly, if you want the entire third column, leave the row spot empty:

```{r}
mat[ ,3]
```

This is also a vector, not a matrix.

You can access more than one column or more than one row if you like. This will give you a new matrix.

```{r}
mat[ ,2:4]
```

You can subset both rows and columns:

```{r}
mat[1:2,2:4]
```

In some of our illustrations, we will convert matrices into data frames so that we can, for example, use them with dplyr. For this, we can use the function `as.data.frame`:

```{r}
as.data.frame(mat)
```

You can also use the `[` to access rows and columns of a data frame:

```{r}
library(dslabs)
data(murders)
murders[25, 1]
murders[2:3, ]
```
