# Sorting

Now that we have mastered some basic R knowledge, let's try to gain some insights into the safety of different states in the context of gun murders. 

## `sort` 

Say we want to rank the states from least to most gun murders. The function `sort` sorts a vector in increasing order. We can therefore see the largest number of gun murders by typing:

```{r}
library(dslabs)
data(murders)
sort(murders$total)
```

However, this does not give us information about which states have which murder totals. For example, we don't know which state had `r max(murders$total)`.

## `order`

The function `order` is closer to what we want. It takes a vector as input and returns the vector of indexes that sorts the input vector. This may sound confusing so let's look at a simple example. We can create a vector and sort it:

```{r}
x <- c(31, 4, 15, 92, 65)
sort(x)
```

Rather than sort the input vector, the function `order` returns the index that sorts input vector:

```{r}
index <- order(x)
x[index]
```

This is the same output as that returned by `sort(x)`. If we look at this index, we see why it works:
```{r}
x
order(x)
```

The second entry of `x` is the smallest, so `order(x)` starts with `2`. The next smallest is the third entry, so the second entry is `3` and so on. 

How does this help us order the states by murders? First, remember that the entries of vectors you access with `$` follow the same order as the rows in the table. For example, these two vectors containing state names and abbreviations respectively are matched by their order:

```{r}
murders$state[1:10]
murders$abb[1:10]
```

This means we can order the state names by their total murders. We first obtain the index that orders the vectors according to murder totals and then index the state names vector:

```{r}
ind <- order(murders$total) 
murders$abb[ind] 
```

According to the above, California had the most murders.

## `max` and `which.max`

If we are only interested in the entry with the largest value, we can use `max` for the value:

```{r}
max(murders$total)
```

and `which.max` for the index of the largest value:

```{r}
i_max <- which.max(murders$total)
murders$state[i_max]
```

For the minimum, we can use `min` and `which.min` in the same way.

Does this mean California the most dangerous state? In an upcoming section, we argue that we should be considering rates not totals. Before doing that, we introduce one last order related function: `rank`

## `rank`

Although not as frequently used as `order` and `sort`, the function `rank` is also related to order and can be useful.
For any given vector it returns a vector with the rank of the first entry, second entry, etc., of the input vector. Here is a simple example:

```{r}
x <- c(31, 4, 15, 92, 65)
rank(x)
```

To summarize, let's look at the results of the three functions we have introduced:

```{r, echo=FALSE}
knitr::kable(data.frame(original=x, sort=sort(x), order=order(x), rank=rank(x)))

```


## Beware of recycling

Another common source of unnoticed errors in R is the use of _recycling_. We saw that vectors are added elementwise. So if the vectors don't match in length, it is natural to assume that we should get an error. But we don't. Notice what happens:

```{r, warning=TRUE}
x <- c(1,2,3)
y <- c(10, 20, 30, 40, 50, 60, 70)
x+y
```
We do get a warning but no error. For the output, R has recycled the numbers in `x`. Notice the last digit of numbers in the output. 
 

## Exercise 

For these exercises we will use the US murders dataset. Make sure you load it prior to starting.

```{r}
library(dslabs)
data("murders")
```

1. Use the `$` operator to access the population size data and store it the object `pop`. Then use the `sort` function to redefine `pop` so that it is sorted. Finally, use the `[` operator to report the smallest population size.


2. Now instead of the smallest population size, find the index of the entry with the smallest population size. Hint: use `order` instead of `sort`.
  
3. We can actually perform the same operation as in the previous exercise using the function `which.min`. Write one line of code that does this.

  
4. Now we know how small the smallest state is and we know which row represents it. Which state is it? Define a variable `states` to be the state names from the `murders` data frame. Report the name of the state with the smallest population.

  
5. You can create a data frame using the `data.frame` function. Here is a quick example:


    ```{r}
    temp <- c(35, 88, 42, 84, 81, 30)
    city <- c("Beijing", "Lagos", "Paris", "Rio de Janeiro", "San Juan", "Toronto")
    city_temps <- data.frame(name = city, temperature = temp)
    ```

    Use the `rank` function to determine the population size rank (from smallest to biggest) of each state. Save these ranks in an object called `ranks`, then create a data frame with the state name and its rank. Call the data frame `my_df`.

   
6. Repeat the previous exercise, but this time order `my_df` so that the states are ordered from least populous to most populous. Hint: create an object `ind` that stores the indexes needed to order the population values. Then use the bracket operator `[` to re-order each column in the data frame.

  
7. The `na_example` represents a series of counts. You can quickly examine the object using:
    
    ```{r}
    data("na_example")  
    str(na_example)
    ```
    However, when we compute the average with the function `mean`, we obtain an `NA`:
    
    
    ```{r}
    mean(na_example)
    ```

    The `is.na` returns a logical vector that tells us which entries are `NA`. Assign this logical vector to an object called `ind` and determine how many `NA`s does `na_example` have.
    
   
8. Now compute the average again, but only for the entries that are not `NA`. Hint: remember the `!` operator.

 
