# The tidyverse {#tidyverse}

We have learned about the __dplyr__, __readr__, and __ggplot2__ packages and will be using them extensively throughout the book. These are part of a collection of packages referred to as the _tidyverse_. The name comes from the fact that these packages work best when the data is in _tidy_ format, a concept we define in the first section of this chapter. 

We can load all the tidyverse packages at once by installing and loading the __tidyverse__ package:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

The approach that makes the tidyverse attractive is that most function take a tidy data frame as input and return a tidy data frame as output. This permits the data analysts to focus on more important aspects of the analysis rather than the format of the data. We will be learning how to implement the tidyverse approach throughout the book. In this chapter we provide a general introduction to the tidyverse approach and a reference guide to some of the functions we have not yet covered and are widely used throughout the book.

## Tidy data {#tidy-data}

To help define tidy data, we will create a table called `tidy_data` using fertility data provided by the `gapminder` dataset. We introduce this dataset in detail in the Data Visualization part of the book. For illustrative purposes we filter the table to include just two countries. 

```{r}
library(dslabs)
data("gapminder")
tidy_data <- gapminder %>% 
  filter(country %in% c("South Korea", "Germany") & !is.na(fertility)) %>%
  select(country, year, fertility)
head(tidy_data)
```

With the data in this format, we could quickly make the desired plot.

```{r tidy-example}
tidy_data %>% 
  ggplot(aes(year, fertility, color = country)) +
  geom_point()
```

One reason this code works seamlessly is because each point is represented in a row. This brings us to the definition of _tidy data_: each row represents one observation and the columns represent the different variables available for each of these observations.

If we go back to the original data from which the `gapmider` dataset was obtained, we see that it does not start out _tidy_. We include an example file with the data shown in this graph mimicking the way it was originally saved in a spreadsheet:

```{r, message=FALSE}
path <- system.file("extdata", package="dslabs")
filename <- file.path(path, "fertility-two-countries-example.csv")
wide_data <- read_csv(filename)
```

The object `wide_data` includes the same information as the object `tidy_data` except it's in a different format, a `wide` format. Here are the first nine columns:

```{r}
select(wide_data, country, `1960`:`1967`)
```

Two important differences between the wide and tidy formats are: 1) in the wide format, each row includes several observations and 2) one of the variables, year, is stored in the header.
 
The __ggplot2__ code we introduced earlier does not work with `wide_data`. For one, there is no `year` variable. To use the __tidyverse__, we need to wrangle this data into `tidy` format, which you will learn to do in the Data Wrangling part of the book.

## Exercises 


1. Examine the dataset `co2`. Which of the following is true:

    A. `co2` is tidy data: it has one year for each row.

    B. `co2` is not tidy: we need at least one column with a character vector.

    C. `co2` is not tidy:  it is a matrix not a data frame.

    D. `co2` is not tidy: to be tidy we would have to wrangle it to have three columns: year, month and value; then each co2 observation has a row.


2. Examine the dataset  `ChickWeight`. Which of the following is true:

    A.  `ChickWeight` is not tidy: each chick has more than one row. 

    B.  `ChickWeight` is tidy: each observation, here a weight, is represented by one row. The chick from which this measurement came from is one the variables.

    C.  `ChickWeight` is not a tidy: we are missing the year column.

    D.  `ChickWeight` is tidy: it is stored in a data frame.


3. Examine the dataset `BOD`. Which of the following is true:

    A. `BOD` is not tidy: it only has six rows.

    B. `BOD` is not tidy: the first column is just an index.

    C. `BOD` is tidy: each row is an observation with two values, time and demand.

    D. `BOD` is tidy: all small datasets are tidy by definition.


4. Which of the following datasets is tidy (you can pick more than one):

    A. `BJsales`

    B. `EuStockMarkets`

    C. `DNase`

    D. `Formaldehyde`

    E. `Orange`

    F. `UCBAdmissions`



## Tibbles

Tidy data must be stored in data frames. We introduced the data frame in Section \@ref(data-frames) and have been using the `murders` data frame throughout the book:

```{r}
data(murders)
class(murders)
```

In Section \@ref(group-by) we introduced the `group_by` function, which permits stratifying data before computing summary statistics. But where is the group information stored in the data.frame? 

```{r}
murders %>% group_by(region) %>% head()
```

Notice that there are no columns with this information. But, if you look closely at the output above, you see the line `A tibble: 6 x 5`. We can learn the class of the returned object using:

```{r}
murders %>% group_by(region)%>% class()
```

The `tbl`, pronounced tibble, is a special kind of data frame. The functions `group_by` and `summarize` always return this type of data frame. The `group_by` function returns a special kind of `tbl`, the `grouped_df`. We will say more about these later. For consistency, the manipulation verbs, `select`, `filter`, `mutate`, and `arrange`, preserve the class of the input: if they receive a regular data frame, they return a regular data frame. If they receive a tibble they return a tibble. But tibbles are the preferred format in the tidyverse and as result parsers that produce a data frame from scratch,
 such as `read_csv` or `read_excel`, return a tibble.

Tibbles are very similar to data frames. In fact, you can think of them as a modern version of data frames. Nonetheless there are three important differences which we describe in the next section.


### Tibbles display better 

The print method for tibbles is more readable than that of a data frame. To see this, compare the outputs of typing `murders` and the output of murders if we convert it to a tibble. We can do this using `as_tibble(murders)`. If using RStudio, output for a tibble adjusts to your window size. To see this, change the width of your R console and notice how more/less columns are shown.

### Subsets of tibbles are tibbles

If you subset the columns of a data frame, you may get back an object that is not a data frame. For example:

```{r}
class(murders[,4])
```

is not a data frame. With tibbles this does not happen:

```{r}
class(as_tibble(murders)[,4])
```

This is useful in the tidyverse since functions require data frames as input. 

With tibbles, if you want to access the vector that defines a column, and not get back a data frame, you need to use the accessor `$`:

```{r}
class(as_tibble(murders)$population)
```

A related feature is that tibbles will give you a warning if you try to access a column that does not exist. If we accidentally write `Population` instead of `population` this:

```{r}
murders$Population
```

returns a `NULL` with no warning, which can make it harder to debug. In contrast, if we try this with a tibble we get an informative warning:

```{r}
as_tibble(murders)$Population
```


### Tibbles can have complex entries

While data frame columns need to be vectors of numbers, strings or Boolean, tibbles can have more complex objects, such as lists or functions. Also, we can create tibbles  with functions:

```{r}
tibble(id = c(1, 2, 3), func = c(mean, median, sd))
```


### Tibbles can be grouped

The function `group_by` returns a special kind of tibble: a grouped tibble. This class stores information that lets you know which rows are in which groups. The tidyverse functions, in particular the `summarize` function, are aware of the group information. 

### Creating a tibble

It is sometimes useful for us to create our own data frames. To create a data frame in the tibble format, you can do this by using the `data_frame` function.   

```{r}
grades <- data_frame(names = c("John", "Juan", "Jean", "Yao"), 
                     exam_1 = c(95, 80, 90, 85), 
                     exam_2 = c(90, 85, 85, 90))
grades
```

Note that R based has a function with a very similar name, `data.frame`, that can be used to create a regular data.frame.  One important difference is that by default `data.frame` coerces characters into factors without providing a warning or message.

```{r}
grades <- data_frame(names = c("John", "Juan", "Jean", "Yao"), 
                     exam_1 = c(95, 80, 90, 85), 
                     exam_2 = c(90, 85, 85, 90))
class(grades$names)
```

To avoid this, we use the rather cumbersome argument `stringsAsFactors`:
```{r}
grades <- data.frame(names = c("John", "Juan", "Jean", "Yao"), 
                     exam_1 = c(95, 80, 90, 85), 
                     exam_2 = c(90, 85, 85, 90),
                     stringsAsFactors = FALSE)
class(grades$names)
```

To covert a regular data frame to a tibble you can use the  `as_tibble` function. They are equivalent. 

```{r}
as_tibble(grades) %>% class()
```


## The dot operator

One of the advantages of using the pipe `%>%` is that we do not have to keep naming new objects as we manipulate the data frame. As quick reminder, if we want to compute the median murder rate for states in the southern states, instead of typing

```{r}
tab_1 <- filter(murders, region == "South")
tab_2 <- mutate(tab_1, rate = total / population * 10^5)
rates <- tab_2$rate
median(rates)
```

We can avoid defining any new objects by instead typing

```{r}
filter(murders, region == "South") %>% 
  mutate(rate = total / population * 10^5) %>% 
  summarize(median = median(rate)) %>%
  pull(median)
```

The reason we can do this is because each of these functions takes a data frame as the first argument. But what if we want to access a component of the data frame. For example, what if the `pull` function was not available and we wanted to access `tab_2$rate`? What data frame name would we use? The answer is the dot operator.

For example to access the rate vector without the `pull` function we could use

```{r}
rates <-filter(murders, region == "South") %>% 
  mutate(rate = total / population * 10^5) %>% 
  .$rate
median(rates)
```

In the next section, we will see other instances in which using the `.` is useful. 

## `do` {#do}

The tidyverse functions know how to interpret grouped tibbles. Furthermore, to facilitate stringing commands through the pipe `%>%`, tidyverse functions consistently return data frames, since this assures that the output of a function is accepted as the input of another. But most R functions do not recognize grouped tibbles nor do they return data frames. The `quantile` function is an example we described in Section \@ref(summarize). The `do` functions serves as a bridge between R functions, such as `quantile` and the tidyverse. The `do` function understands grouped tibbles and always returns a data frame.

In Section \@ref(summarize) we noted that if you attempt to use `quantile` to obtain the min, median and max in one call we will receive an error: `Error: expecting result of length one, got : 2`.

```{r, eval=FALSE}
data(heights)
heights %>% 
  filter(sex == "Female") %>%
  summarize(range = quantile(height, c(0, 0.5, 1)))
```

We can use the `do` function fix this. 

First we have to write a function that fits into the tidyverse approach, receives a data frame and returns a data frame. 

```{r}
my_summary <- function(dat){
  x <- quantile(dat$height, c(0, 0.5, 1))
  data_frame(min = x[1], median = x[2], max = x[3])
}
```

We can now apply the function to the heights dataset to obtain the summaries:

```{r}
heights %>% 
  group_by(sex) %>% 
  my_summary
```

But this is not what we want. We want a summary for each sex and the code returned just one summary. This is because `my_summary` is not part of the tidyverse and does not know how to handled grouped tibbles. `do` makes this connection:

```{r}
heights %>% 
  group_by(sex) %>% 
  do(my_summary(.))
```

Note that here we need to use the dot operator. The tibble created by `group_by` is the data frame that is pipped to `do`. Within the call to `do`, the name of this data frame is `.` and we want to send it to `my_summary`. If you do not use the dot, then `my_summary` has no argument and returns an error telling us that `argument "dat"` is missing. You can see the error by typing:

```{r, eval=FALSE}
heights %>% 
  group_by(sex) %>% 
  do(my_summary())
```

If you do not use the parenthesis, then the function is not executed and instead do tries to return the function. This gives an error because do must always return a data frame. You can see the error by typing:


```{r, eval=FALSE}
heights %>% 
  group_by(sex) %>% 
  do(my_summary)
```

## The purrr package

In Section \@ref(vectorization) we learned about the `sapply` function, which permitted us to apply the same function to each element of a vector. We constructed this function

```{r}
compute_s_n <- function(n){
  x <- 1:n
  sum(x)
}
```

and used `sapply` to compute the sum of the first `n` integers for several values of `n` like this:

```{r}
n <- 1:25
s_n <- sapply(n, compute_s_n)
```

This type of operation, applying the same function or procedure to elements of an object is quite common in data analysis. The __purrr__ package includes functions like `sapply` but that better interact with other tidyverse functions. The main advantage is that we can better control the output of functions.
In contrast `sapply` can return several different object types. 

The first __purrr__ function we will learn is `map`, which works very similar to `sapply`, but always, without exception, returns a list. 

The main __purrr__ function is `map` 

```{r}
library(purrr)
s_n <- map(n, compute_s_n)
class(s_n)
```

If we want a numeric vector, we can instead use `map_dbl` which always returns a vector of numeric values. 

```{r}
s_n <- map_dbl(n, compute_s_n)
class(s_n)
```

This produces the same results as the `sapply` call shown above.

A particularly useful __purrr__ function for interacting with the rest of the tidyverse is `map_df` which always returns a tibble data frame. However, the function being called needs to return a vector or list with names. For this reason, the following code would result in a `Argument 1 must have names` error.

```{r, eval=FALSE}
s_n <- map_df(n, compute_s_n)
```

We need to change the function to make this work:

```{r}
compute_s_n <- function(n){
  x <- 1:n
  data_frame(sum = sum(x))
}
s_n <- map_df(n, compute_s_n)
head(s_n)
```
The __purrr__ package provides much more functionality not covered here. For more details you can consult [this online resource](https://jennybc.github.io/purrr-tutorial/).

## Conditionals

A typical data analysis will often involve one or more conditional operation. In Section \@ref(conditionals) we described the `ifelse` function, which we will use extensively in this book. In this section we present two __dplyr__ functions that provide further functionality for performing conditional operations.

### `case_when`

The `case_when` function is useful for vectorizing conditional statements. It is similar to `ifelse` but can output any number of values, as opposed to just `TRUE` or `FALSE`. Here is an example splitting numbers into negative, positives and 0:

```{r}
x <- c(-2, -1, 0, 1, 2)
case_when(x < 0 ~ "Negative", x > 0 ~ "Positive", TRUE ~ "Zero")
```

A common use for this function is to define categorical variables based on existing variables. So, for example, suppose we we want compare the murder rates in in three groups of states: _New England_, _West Coast_, _South_,  and _other_. For each state, we need to ask if it is in New England, if it is not we ask if it is in the West Coast, if not we ask if it is in the South and if not we assign _other_. Here is how we use `case_when` to do this:

```{r}
data(murders)
murders %>% 
  mutate(group = case_when(
    abb %in% c("ME", "NH", "VT", "MA", "RI", "CT") ~ "New England",
    abb %in% c("WA", "OR", "CA") ~ "West Coast",
    region == "South" ~ "South",
    TRUE ~ "other")) %>%
  group_by(group) %>%
  summarize(rate = sum(total) / sum(population) * 10^5) %>%
  arrange(rate)
```

### `between`

A common operation in data analysis is to determine if a value falls inside an interval. We can check this using conditionals. For example to check if the elements of a vector `x` are between `a` and `b` we can type

```{r, eval=FALSE}
x >= a & x <= b
```

However, this can become cumbersome, especially within the tidyverse approach. The  `between` function performs the same operation.

```{r, eval = FALSE}
between(x, a, b)
```

## Exercises 

1. Load the `murders` dataset. Which of the following is true?

    A. `murders` is in tidy format and is stored in a tibble
    
    B. `murders` is in tidy format and is stored in a data frame
    
    C. `murders` is not in tidy format and is stored in a tibble
    
    D. `murders` is not in tidy format and is stored in a data frame.
    
2. Use the `as_tibble` to covert the `muders` data table into a tibble and save it in an object called `murders_tibble`.
    
3. Use the `group_by` function to cover murders into a tibble that is groups by region.

4. Write tidyverse code that is equivalent to this code. `exp(mean(log(murders$population)))`. Write it using the pipe so that each function is called without arguments. Use the dot operator to access the population. Hint: The code should start with `murders %>%`.
    
5. Use the `map_df` to create a data frame with three columns with names `n`, `s_n`, and `s_n_2`. The first column should contain the numbers 1 through 100. The second and third columns should contain the sum of 1 through $n$ with $n$ the row number.  
