# Vectors

In R, the most basic objects available to store data are _vectors_. As we have seen, complex datasets can usually be broken down into components that are vectors. For example, in a data frame, each column is a vector. Here we learn more about this important class.

## Creating vectors

We can create vectors using the function `c`, which stands for _concatenate_. We use `c` to concatenate entries in the following way:

```{r}
codes <- c(380, 124, 818)
codes
```

We can also create character vectors. We use the quotes to denote that the entries are characters rather than variable names.

```{r}
country <- c("italy", "canada", "egypt")
```

In R you can also use single quotes:

```{r}
country <- c('italy', 'canada', 'egypt')
```

But be careful not to confuse the single quote ' with the _back quote_ `. 

By now you should know that if you type:

```{r, eval=FALSE}
country <- c(italy, canada, egypt)
```
you receive an error because the variables `italy`, `canada` and `egypt` are not defined. If we do not use the quotes, R looks for variables with those names and returns an error.

## Names

Sometimes it is useful to name the entries of a vector. For example, when defining a vector of country codes, we can use the names to connect the two:

```{r}
codes <- c(italy = 380, canada = 124, egypt = 818)
codes
```

The object `codes` continues to be a numeric vector:
```{r}
class(codes)
```
but with names:
```{r}
names(codes)
```

If the use of strings without quotes looks confusing, know that you can use the quotes as well:

```{r}
codes <- c("italy" = 380, "canada" = 124, "egypt" = 818)
codes
```

There is no difference between this function call and the previous one. This is one of the many ways in which R is quirky compared to other languages.

We can also assign names using the `names` functions:

```{r}
codes <- c(380, 124, 818)
country <- c("italy","canada","egypt")
names(codes) <- country
codes
```

## Sequences

Another useful function for creating vectors generates sequences:

```{r}
seq(1, 10)
```

The first argument defines the start, and the second defines the end which is included. The default is to go up in increments of 1, but a third argument lets us tell it how much to jump by:

```{r}
seq(1, 10, 2)
```

If we want consecutive integers, we can use the following shorthand:

```{r}
1:10
```

When we use these functions, R produces integers, not numerics, because they are typically used to index something:

```{r}
class(1:10)
```

However, as soon as we create sequence including non-integer, the class changes:

```{r}
class(seq(1, 10, 0.5))
```

## Subsetting

We use square brackets to access specific elements of a vector. For the vector `codes` we defined above, we can access the second element using:
```{r}
codes[2]
```

You can get more than one entry by using a multi-entry vector as an index:
```{r}
codes[c(1,3)]
```

The sequences defined above are particularly useful if we want to access, say, the first two elements:

```{r}
codes[1:2]
```

If the elements have names, we can also access the entries using these names. Below are two examples.

```{r}
codes["canada"]
codes[c("egypt","italy")]
```

## Coercion

In general, _coercion_ is an attempt by R to be flexible with data types. When an entry does not match the expected, some of the prebuilt R functions try to guess what was meant before throwing an error. This can also lead to confusion. Failing to understand _coercion_ can drive programmers crazy when attempting to code in R since it behaves quite differently from most other languages in this regard. Let's learn about it with some examples.

We said that vectors must be all of the same type. So if we try to combine, say, numbers and characters, you might expect an error:

```{r}
x <- c(1, "canada", 3)
```

But we don't get one, not even a warning! What happened? Look at `x` and its class:

```{r}
x
class(x)
```

R _coerced_ the data into characters. It guessed that because you put a character string in the vector, you meant the 1 and 3 to actually be character strings `"1"` and "`3`". The fact that not even a warning is issued is an example of how coercion can cause many unnoticed errors in R. 

R also offers functions to change from one type to another. For example, you can turn numbers into characters with:

```{r}
x <- 1:5
y <- as.character(x)
y
```

You can turn it back with `as.numeric`:

```{r}
as.numeric(y)
```
 
This function is actually quite useful since datasets that include numbers as character strings are common. 

## Not availables (NA)

When a function tries to coerce one type to another and encounters an impossible case, it usually gives us a warning and turns the entry into a special value called an `NA` for "not available".  For example:

```{r}
x <- c("1", "b", "3")
as.numeric(x)
```

R does not have any guesses for what number you want when you type `b`, so it does not try.

As a data scientist, you will encounter the `NA`s often as they are generally used for missing data, a common problem in real-world datasets.

## Exercises

1. Use the function `c` to create a vector with the average high temperatures in January for Beijing, Lagos, Paris, Rio de Janeiro, San Juan, and Toronto, which are 35, 88, 42, 84, 81, and 30 degrees Fahrenheit. Call the object `temp`.

2. Now create a vector with the city names and call the object `city`.


3. Use the `names` function and the objects defined in the previous exercises to associate the temperature data with its corresponding city.

 
4. Use the `[` and `:` operators to access the temperature of the first three cities on the list.
    
 
5. Use the `[` operator to access the temperature of Paris and San Juan. 
   
6. Use the `:` operator to create a sequence of numbers $12,13,14,\dots,73$. 
   
7. Create a vector containing all the positive odd numbers smaller than 100.

8. Create a vector of numbers that starts at 6, does not pass 55, and adds numbers in increments of 4/7: 6, 6+4/7, 6+8/7, etc.. How many numbers does the list have? Hint: use `seq` and `length`.
 
9. What is the class of the following object `a <- seq(1, 10, 0.5)`?



10. What is the class of the following object `a <- seq(1, 10)`?

   
11. The class of `class(a<-1)` is numeric not integer. R defaults to numeric and to force an integer, you need to add the letter `L`. Confirm that the class of `1L` is integer.

  
12. Define the following vector:

    ```{r}
    x <- c("1", "3", "5")
    ```
    and coerce it to get integers.
  
  




