## Bayesian statistics {#bayesian-statistics}

What does it mean when an election forecaster tells us that a given candidate has a 90% chance of winning? In the context of the urn model, this would be equivalent to stating that the probability $p>0.5$ is 90%. However, as we discussed earlier, in the urn model $p$ is a fixed parameter and it does not make sense to talk about probability. With Bayesian statistics, we model $p$ as random variable and thus a statement such as "90% chance of winning" is consistent with the approach.

Forecasters also use models to describe variability at different levels. For example, sampling variability, pollster to pollster variability, day to day variability, and election to election variability. One of the most successful approaches used for this are hierarchical models, which can be explained in the context of Bayesian statistics.

In this chapter we briefly describe Bayesian statistics. For an in-depth treatment of this topic we recommend one of the following textbooks:

* Berger JO (1985). Statistical Decision Theory and Bayesian Analysis, 2nd edition. Springer-Verlag. 

* Lee PM (1989). Bayesian Statistics: An Introduction. Oxford. 


### Bayes theorem

We start by describing Bayes theorem. We do this using a hypothetical cystic fibrosis test as an example. 
Suppose a test for cystic fibrosis has an accuracy of 99%. We will use the following notation:

$$
\mbox{Prob}(+ \mid D=1)=0.99, \mbox{Prob}(- \mid D=0)=0.99 
$$

with $+$ meaning a positive test and $D$ representing if you actually have the disease (1) or not (0).

Suppose we select a random person and they test positive. What is the probability that they have the disease?  We write this as $\mbox{Prob}(D=1 \mid +)?$ The cystic fibrosis rate is 1 in 3,900 which implies that  $\mbox{Prob}(D=1)=0.00025$. To answer this question, we will use Bayes theorem, which in general tells us that:

$$
\mbox{Pr}(A \mid B)  =  \frac{\mbox{Pr}(B \mid A)\mbox{Pr}(A)}{\mbox{Pr}(B)} 
$$

This equation applied to our problem becomes:

$$
\begin{aligned}
\mbox{Pr}(D=1 \mid +) & =  \frac{ P(+ \mid D=1) \cdot P(D=1)} {\mbox{Pr}(+)} \\
& =  \frac{\mbox{Pr}(+ \mid D=1)\cdot P(D=1)} {\mbox{Pr}(+ \mid D=1) \cdot P(D=1) + \mbox{Pr}(+ \mid D=0) \mbox{Pr}( D=0)} 
\end{aligned}
$$

Plugging in the numbers we get:

$$
\frac{0.99 \cdot 0.00025}{0.99 \cdot 0.00025 + 0.01 \cdot (.99975)}  =  0.02 
$$

This says that despite the test having 0.99 accuracy, the probability of having the disease given a positive test is only 0.02. This may appear counter-intuitive to some, but the reason this is the case is because we have to factor in the very rare probability that a person, chosen at random, has the disease. To illustrate this, we run a Monte Carlo simulation.

## Bayes theorem simulation

The following simulation is meant to help you visualize Bayes theorem. We start by randomly selecting 100,000 people from a population in which the disease in question has a 1 in 4,000 prevalence.

```{r, echo=FALSE}
set.seed(3)
```

```{r}
prev <- 0.00025
N <- 100000
outcome <- sample(c("Disease","Healthy"), N, replace = TRUE, 
                  prob = c(prev, 1 - prev))
```

Note that there are very few people with the disease:

```{r}
N_D <- sum(outcome == "Disease")
N_D
N_H <- sum(outcome == "Healthy")
N_H
```

Also, there are many without the disease, which makes it more probable that we will see some false positives given that the test is not perfect. Now each person gets the test, which is correct 99% of the time:

```{r}
accuracy <- 0.99
test <- vector("character", N)
test[outcome == "Disease"]  <- sample(c("+", "-"), N_D, replace = TRUE, 
                                    prob = c(accuracy, 1 - accuracy))
test[outcome == "Healthy"]  <- sample(c("-", "+"), N_H, replace = TRUE, 
                                    prob = c(accuracy, 1 - accuracy))
```

Because there are so many more controls than cases, even with a low false positive rate we get more controls than cases in the group that tested positive:


```{r}
table(outcome, test)
```

From this table, we see that the proportion of positive tests that have the disease is `r sum(test=="+" & outcome=="Disease")` out of `r sum(test=="+")`. We can run this over and over again to see that, in fact, the probability converges to about 0.022.


### Bayes in practice


José Iglesias is a professional baseball player. In April 2013, when he was starting his career, he was performing rather well:

| Month | At Bats | H | AVG |
|-------|---------|---|-----|
| April | 20      | 9 | .450|

The batting average (`AVG`) statistic is one way of measuring success. Roughly speaking, it tells us the success rate when batting. An `AVG` of .450 means José has been successful 45% of the times he has batted (`At Bats`) which is rather high, historically speaking. Keep in mind that no one has finished a season with an `AVG` of .400 or more since Ted Williams did it in 1941! To illustrate the way hierarchical models are powerful, we will try to predict José's batting average at the end of the season. Note that in a typical season, players have about 500 at bats.

With the techniques we have learned up to now, referred to as _frequentist techniques_, the best we can do is provide a confidence interval. We can think of outcomes from hitting as a binomial with a success rate of $p$. So if the success rate is indeed .450, the standard error of just 20 at bats is:

$$
\sqrt{\frac{.450 (1-.450)}{20}}=.111
$$

This means that our confidence interval is $.450 - .222$ to $.450 + .222$ or $.228$ to $.672$.

This prediction has two problems. First, it is very large, so not very useful. Second, it is centered at .450, which implies that our best guess is that this new player will break Ted Williams' record. 

If you follow baseball, this last statement will seem wrong and this is because you are implicitly using a hierarchical model that factors in information from years of following baseball. Here we show how we can quantify this intuition.

First, let's explore the distribution of batting averages for all players with more than 500 at bats during the previous three seasons:

```{r batting-averages-histogram, echo=FALSE, out.width="100%", fig.height=3, message=FALSE, warning=FALSE}
library(tidyverse)
library(Lahman)
filter(Batting, yearID %in% 2010:2012) %>% 
  mutate(AVG = H/AB) %>% 
  filter(AB > 500) %>% 
  ggplot(aes(AVG)) +
  geom_histogram(color="black", binwidth = .01) +
  facet_wrap( ~ yearID)
```

The average player had an `AVG` of .275 and the standard deviation of the population of players was 0.027. So we can see already that .450 would be quite an anomaly since it is over six standard deviations away from the mean. 

So is José lucky or is he the best batter seen in the last 50 years? Perhaps it's a combination of both luck and talent. But how much of each? If we become convinced that he is lucky, we should trade him to a team that trusts the .450 observation and is maybe overestimating his potential.


## Hierarchical models

The hierarchical model provides a mathematical description of how we came to see the observation of .450. First, we pick a player at random with an intrinsic ability summarized by, for example, $p$. Then we see 20 random outcomes with success probability $p$.

We use a model to represent two levels of variability in our data. First, each player is assigned a natural ability to hit. We will use the symbol $p$ to represent this ability. You can think of $p$ as the batting average you would converge to if this particular player batted over and over again. 

Based on the plots we showed earlier, we assume that $p$ has a normal distribution. With expected value .270 and standard error 0.027. 

Now the second level of variability has to do with luck when batting. Regardless of how good the player is, sometimes you have bad luck and sometimes you have good luck. At each at bat, this player has a probability of success $p$. If we add up these successes and failures, then the CLT tells us that the observed average, call it $Y$, has a normal distribution with expected value $p$ and standard error $\sqrt{p(1-p)/N}$ with $N$ the number of at bats.

Statistical textbooks will write the model like this:
$$
\begin{aligned}
p &\sim N(\mu, \tau^2) \\
Y \mid p &\sim N(p, \sigma^2) 
\end{aligned}
$$
Here the $\sim$ symbol tells us the random variable on the left of the symbol follows the distribution on the right and $N(a,b^2)$ represents the normal distribution with mean $a$ and standard deviation $b$. The $\mid$ is read as _conditioned on_, and it means that we are treating the random variable to the right of the symbol as known. We refer to the model as hierarchical because we need to know $p$, the first level, in order to model $Y$, the second level. In our example the first level describes randomness in assigning talent to a player and the second describes randomness in this particular player's performance once we have fixed the talent parameter. In a Bayesian framework, the first level is called a _prior distribution_ and the second the _sampling distribution_. The data analysis we have conducted here suggests that we set $\mu = .270$, $\tau = 0.027$, and $\sigma^2 = p(1-p)/N$.

Now, let's use this model for José's data. Suppose we want to predict his innate ability in the form of his _true_ batting average $p$. This would be the hierarchical model for our data:

$$
\begin{aligned}
p &\sim N(.275, .027^2) \\
Y \mid p &\sim N(p, .111^2) 
\end{aligned}
$$

We now are ready to compute a posterior distribution to summarize our prediction of $p$. The continuous version of Bayes' rule can be used here to derive the _posterior probability function_, which is the distribution of $p$ assuming we observe $Y=y$. In our case, we can show that when we fix $Y=y$, $p$ follows a normal distribution with expected value:

$$
\begin{aligned}
\mbox{E}(p \mid Y=y) &= B \mu + (1-B) y\\
&= \mu + (1-B)(y-\mu)\\
\mbox{with } B &= \frac{\sigma^2}{\sigma^2+\tau^2}
\end{aligned}
$$

This is a weighted average of the population average $\mu$ and the observed data $y$. The weight depends on the SD of the population $\tau$ and the SD of our observed data $\sigma$. This weighted average is sometimes referred to as _shrinking_ because it _shrinks_ estimates towards a prior mean. In the case of José Iglesias, we have:

$$
\begin{aligned}
\mbox{E}(p \mid Y=.450) &= B \times .275 + (1 - B) \times .450 \\
&= .275 + (1 - B)(.450 - .275) \\
B &=\frac{.111^2}{.111^2 + .027^2} = 0.944\\
\mbox{E}(p \mid Y=450) &\approx .285
\end{aligned}
$$

We do not show the derivation here, but the standard error can be shown to be:

$$
\mbox{SE}(p\mid y)^2 = \frac{1}{1/\sigma^2+1/\tau^2}
= \frac{1}{1/.111^2 + 1/.027^2} = 0.00069
$$
and the standard deviation is therefore $0.026$. So we started with a frequentist 95% confidence interval that ignored data from other players and summarized just José's data: .450 $\pm$ 0.220. Then we used a Bayesian approach that incorporated data from other players and other years to obtain a posterior probability. This is actually referred to as an empirical Bayes approach because we used data to construct the prior. From the posterior, we can report what is called a 95% _credible interval_ by reporting a region, centered at the mean, with a 95% chance of occurring. In our case, this turns out to be: .285 $\pm$ 0.052.

The Bayesian credible interval suggests that if another team is impressed by the .450 observation, we should consider trading José as we are predicting he will be just slightly above average. Interestingly, the Red Sox traded José to the Detroit Tigers in July. Here are the José Iglesias batting averages for the next five months: 

|Month|At Bat| Hits| AVG |
|-----|------|-----|-----|
|April|20|9|.450|
|May|26|11|.423|
|June|86|34|.395|
|July|83|17|.205|
|August|85|25|.294|
|September|50|10|.200|
|Total w/o April|330|97|.293|

Although both intervals included the final batting average, the Bayesian credible interval provided a much more precise prediction. In particular, it predicted that he would not be as good during the remainder of the season.

## Exercises

1\. In 1999, in England, Sally Clark^[https://en.wikipedia.org/wiki/Sally_Clark] was found guilty of the murder of two of her sons. Both infants were found dead in the morning, one in 1996 and another in 1998. In both cases, she claimed the cause of death was sudden infant death syndrome (SIDS). No evidence of physical harm was found on the two infants so the main piece of evidence against her was the testimony of Professor Sir Roy Meadow, who testified that the chances of two infants dying of SIDS was 1 in 73 million. He arrived at this figure by finding that the rate of SIDS was 1 in 8,500 and then calculating that the chance of two SIDS cases was 8,500 $\times$ 8,500 $\approx$ 73 million. Which of the following do you agree with?
    
a. Sir Meadow assumed that the probability of the second son being affected by SIDS was independent of the first son being affected, thereby ignoring possible genetic causes. If genetics plays a role then: $\mbox{Pr}(\mbox{second case of SIDS} \mid \mbox{first case of SIDS}) < \mbox{P}r(\mbox{first case of SIDS})$.
b. Nothing. The multiplication rule always applies in this way: $\mbox{Pr}(A \mbox{ and } B) =\mbox{Pr}(A)\mbox{Pr}(B)$
c. Sir Meadow is an expert and we should trust his calculations.
d. Numbers don't lie.
    
    
2\. Let's assume that there is in fact a genetic component to SIDS and the probability of $\mbox{Pr}(\mbox{second case of SIDS} \mid \mbox{first case of SIDS}) = 1/100$, is much higher than 1 in 8,500. What is the probability of both of her sons dying of SIDS?

3\. Many press reports stated that the expert claimed the probability of Sally Clark being innocent as 1 in 73 million. Perhaps the jury and judge also interpreted the testimony this way. This probability can be written as the probability of _a mother is a son-murdering psychopath_ given that 
_two of her children are found dead with no evidence of physical harm_. 
According to Bayes' rule, what is this?

4\. Assume that the chance of a son-murdering psychopath finding a way to kill her children, without leaving evidence of physical harm, is:

$$
\mbox{Pr}(A \mid B) = 0.50
$$ 

with A = two of her children are found dead with no evidence of physical harm and B = a mother is a son-murdering psychopath = 0.50. Assume that the rate of son-murdering psychopaths mothers is 1 in 1,000,000.     According to Bayes' theorem, what is the probability of $\mbox{Pr}(B \mid A)$ ?


5/. After Sally Clark was found guilty, the Royal Statistical Society issued a statement saying that there was "no statistical basis" for the expert's claim. They expressed concern at the "misuse of statistics in the courts". Eventually, Sally Clark was acquitted in June 2003. What did the expert miss?
    
a. He made an arithmetic error.
b. He made two mistakes. First, he misused the multiplication rule and did not take into account how rare it is for a mother to murder her children. After using Bayes' rule, we found a probability closer to 0.5 than 1 in 73 million.
c. He mixed up the numerator and denominator of Bayes' rule.
d. He did not use R.
    
6\. Florida is one of the most closely watched states in the U.S. election because it has many electoral votes, and the election is generally close, and Florida tends to be a swing state that can vote either way. Create the following table with the polls taken during the last two weeks:

```{r, eval=FALSE}
library(tidyverse)
library(dslabs)
data(polls_us_election_2016)
polls <- polls_us_election_2016 %>% 
  filter(state == "Florida" & enddate >= "2016-11-04" ) %>% 
  mutate(spread = rawpoll_clinton/100 - rawpoll_trump/100)
```

Take the average spread of these polls. The CLT tells us this average is approximately normal. Calculate an average and provide an estimate of the standard error. Save your results in an object called `results`.

7\. Now assume a Bayesian model that sets the prior distribution for Florida's election night spread $d$ to be Normal with expected value $\mu$ and standard deviation $\tau$. What are the interpretations of $\mu$ and $\tau$?
    
a. $\mu$ and $\tau$ are arbitrary numbers that let us make probability statements about $d$.
b. $\mu$ and $\tau$ summarize what we would predict for Florida before seeing any polls. Based on past elections, we would set $\mu$ close to 0 because both Republicans and Democrats have won, and $\tau$ at about $0.02$, because these elections tend to be close.
c. $\mu$ and $\tau$ summarize what we want to be true. We therefore set $\mu$ at $0.10$ and $\tau$ at $0.01$. 
d. The choice of prior has no effect on Bayesian analysis.
    
    
8\. The CLT tells us that our estimate of the spread $\hat{d}$ has normal distribution with expected value $d$ and standard deviation $\sigma$ calculated in problem 6. Use the formulas we showed for the posterior distribution to calculate the expected value of the posterior distribution if we set $\mu = 0$ and $\tau = 0.01$.


9\. Now compute the standard deviation of the posterior distribution.


10\. Using the fact that the posterior distribution is normal, create an interval that has a 95% probability of occurring centered at the posterior expected value. Note that we call these credible intervals.


11\. According to this analysis, what was the probability that Trump wins Florida?

12\. Now use `sapply` function to change the prior variance from `seq(0.05, 0.05, len = 100)` and observe how the probability changes by making a plot.


